### 2020-09-07 Manuscript.Rmd
Initial commit for improved/simplified thesis manuscript repo

---
title: "Missing The Point"
author: "Hanne Oberman"
date: "7-9-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Missing The Point

## Intro

- Missing data has to be accomodated, usually filled in

- Filling in is algorithmic, which relies on convergence

- Diagnosing convergence is difficult, both visually and diagnostically

- In this paper we explore diagnostic methods for MICE, and show which parameter has the best performance

## Identifying non-convergence

- What does non-convergence look like? What are the consequences?

-

## Simulation

- 3x3 missingness conditions: MCAR, MAR, MNAR + 5, 50, 95% of cases incomplete

- outcomes: estimated regression coefficient, CIW, coverage, R2

- convergence parameters: chain means, chain variances, scientific parameter, lambda

- diagnostics: rhat, ac

### 2020-09-07 R/impute.R
Create simulation setup and basic imputation functions

@@ -1,0 +1,38 @@
+# Impute the data
+impute <- function(amp, m_mech, p_inc, it_nr, imp_nr = 5){
+mids <- mice(amp[[m_mech]][[p_inc]], 
+          m = imp_nr,
+          method = "norm",
+          maxit = it_nr, 
+          printFlag = FALSE)
+
+mipo <- mids %>% 
+  mice::lm.mids(Y ~ ., .) %>% 
+  mice::pool() %>% 
+  .$pooled
+
+est <- mipo$estimate[2]
+SE <- sqrt(mipo$b + (mipo$b / mipo$m)) %>% .[2]
+ci_lo <- est - qt(.975, df = imp_nr - 1) * SE 
+ci_up <- est + qt(.975, df = imp_nr - 1) * SE 
+
+pca <- mids %>% mice::complete("all") %>% 
+  purrr::map_dbl(., ~ {
+    princomp(., cor = TRUE) %>% .$sdev %>% .[1] %>% . ^ 2 #first eigenvalue of the varcovar matrix
+  })
+
+out <- data.frame(
+  mm = m_mech,
+  p = p_inc,
+  it = it_nr,
+  est = est,
+  CIW = ci_up - ci_lo,
+  cov = ci_lo < Q & Q < ci_up,
+  pca = t(pca)
+)
+
+return(out)
+}
+
+# test
+impute(amp = amp, m_mech = "MCAR", p_inc = 0.5, it_nr = 1, imp_nr = 5)

### 2020-09-07 R/setup.R

@@ -1,0 +1,28 @@
+# Setup simulation
+setup <- function(seed){
+  # setup environment
+  library(mice)
+  library(tidyverse)
+  set.seed(seed)
+  
+  # create complete data
+  vcov <- matrix(c(1,.5,.5,.5,1,.5,.5,.5,1), ncol = 3)
+  dat <<- mvtnorm::rmvnorm(1000, sigma = vcov) %>% 
+    as.data.frame() %>% 
+    setNames(c("Y", "X1", "X2")) %>% 
+    mutate(Y = Y + rnorm(1000, sd = 1))
+  Q <<- lm(Y~X1+X2, dat) %>% 
+    broom::tidy() %>% 
+    .[["estimate"]] %>% 
+    .[2]
+  
+  # create patterns to ampute the data with multivariate missingness
+  amp_pat <<-
+    expand.grid(c(0, 1), c(0, 1), c(0, 1)) %>% #define all possible combinations of univariate and multivariate missingness
+    .[c(-1,-8), ]  %>% #remove the completely (un)observed cases  
+    setNames(names(ampute(dat)$patterns)) #obtain correct names of patterns
+  
+  # simulation parameters
+  p_inc <<- c(0.05, 0.5, 0.95)
+  m_mech <<- c("MCAR", "MAR", "MCAR")
+}

### 2020-09-07 RunSimulation.R

@@ -1,0 +1,17 @@
+# Improved simulation for thesis manuscript
+
+# setup simulation environment
+source("R/setup.R")
+setup(seed = 1111)
+
+# ampute the data
+amp <- map(m_mech, function(mm) {
+  map(p_inc, function(pp) {
+  mice::ampute(
+    data = dat,
+    prop = pp,
+    patterns = amp_pat,
+    mech = "MCAR"
+  )$amp
+})%>% set_names(as.character(p_inc))}) %>% set_names(as.character(m_mech))
+

### 2020-09-07 R/impute.R
Complete a single simulation run, without chain means and chain variances

@@ -1,38 +1,88 @@
 # Impute the data
-impute <- function(amp, m_mech, p_inc, it_nr, imp_nr = 5){
-mids <- mice(amp[[m_mech]][[p_inc]], 
-          m = imp_nr,
-          method = "norm",
-          maxit = it_nr, 
-          printFlag = FALSE)
+impute <-
+  function(amp,
+           mids,
+           m_mech,
+           p_inc,
+           it_nr,
+           it_total = 50,
+           n_imp = 5) {
+    
+    # initialize the mids object in the first iteration
+    if (it_nr == 1) {
+      mids <<- mice(
+        amp[[m_mech]][[p_inc]],
+        m = n_imp,
+        method = "norm",
+        maxit = 1,
+        printFlag = FALSE
+      )
+    }
+    
+    # append to the mids object in any further iteration
+    if (it_nr > 1) {
+      mids <<- mice.mids(mids, printFlag = FALSE)
+    }
+    
+    # analyze and pool imputations
+    mipo <- mids %>%
+      mice::lm.mids(Y ~ ., .) %>%
+      mice::pool() %>%
+      .$pooled
+    
+    # extract statistics
+    est <- mipo$estimate[2]
+    SE <- sqrt(mipo$b + (mipo$b / n_imp)) %>% .[2]
+    ci_lo <- est - qt(.975, df = n_imp - 1) * SE
+    ci_up <- est + qt(.975, df = n_imp - 1) * SE
+    
+    # add convergence parameters
+    mild <- mids %>% mice::complete("all")
+    qhat <-
+      map_dbl(mild, ~ {
+        lm(formula = Y ~ ., data = .) %>% .$coefficients %>% .[2]
+      })
+    lambda <- mild %>%
+      purrr::map_dbl(., ~ {
+        princomp(., cor = TRUE) %>% .$sdev %>% .[1] %>% . ^ 2 #first eigenvalue of the varcovar matrix
+      })
+    
+    # save for each simulation condition
+    out <- data.frame(
+      mech = m_mech,
+      p = p_inc,
+      it = it_nr,
+      est = est,
+      CIW = ci_up - ci_lo,
+      cov = ci_lo < Q & Q < ci_up,
+      qhat = t(qhat),
+      lambda = t(lambda)
+    )
+    
[…]

### 2020-09-07 R/setup.R

@@ -1,28 +1,31 @@
 # Setup simulation
-setup <- function(seed){
+setup <- function(seed) {
   # setup environment
   library(mice)
   library(tidyverse)
   set.seed(seed)
   
   # create complete data
-  vcov <- matrix(c(1,.5,.5,.5,1,.5,.5,.5,1), ncol = 3)
-  dat <<- mvtnorm::rmvnorm(1000, sigma = vcov) %>% 
-    as.data.frame() %>% 
-    setNames(c("Y", "X1", "X2")) %>% 
+  vcov <- matrix(c(1, .5, .5, .5, 1, .5, .5, .5, 1), ncol = 3)
+  dat <<- mvtnorm::rmvnorm(1000, sigma = vcov) %>%
+    as.data.frame() %>%
+    setNames(c("Y", "X1", "X2")) %>%
     mutate(Y = Y + rnorm(1000, sd = 1))
-  Q <<- lm(Y~X1+X2, dat) %>% 
-    broom::tidy() %>% 
-    .[["estimate"]] %>% 
+  Q <<- lm(Y ~ X1 + X2, dat) %>%
+    broom::tidy() %>%
+    .[["estimate"]] %>%
     .[2]
   
   # create patterns to ampute the data with multivariate missingness
   amp_pat <<-
     expand.grid(c(0, 1), c(0, 1), c(0, 1)) %>% #define all possible combinations of univariate and multivariate missingness
-    .[c(-1,-8), ]  %>% #remove the completely (un)observed cases  
+    .[c(-1, -8),]  %>% #remove the completely (un)observed cases
     setNames(names(ampute(dat)$patterns)) #obtain correct names of patterns
   
   # simulation parameters
   p_inc <<- c(0.05, 0.5, 0.95)
   m_mech <<- c("MCAR", "MAR", "MCAR")
+  
+  # object to store imputations in
+  mids <<- NULL
 }

### 2020-09-07 RunSimulation.R

@@ -7,11 +7,22 @@ setup(seed = 1111)
 # ampute the data
 amp <- map(m_mech, function(mm) {
   map(p_inc, function(pp) {
-  mice::ampute(
-    data = dat,
-    prop = pp,
-    patterns = amp_pat,
-    mech = "MCAR"
-  )$amp
-})%>% set_names(as.character(p_inc))}) %>% set_names(as.character(m_mech))
+    mice::ampute(
+      data = dat,
+      prop = pp,
+      patterns = amp_pat,
+      mech = "MCAR"
+    )$amp
+  }) %>% set_names(as.character(p_inc))
+}) %>% set_names(as.character(m_mech))
 
+results <- map_dfr(m_mech, function(mm) {
+  map_dfr(p_inc, function(pp) {
+    map_dfr(1:50, function(it) {
+  impute(amp,
+  mids,
+  m_mech = mm,
+  p_inc = pp,
+  it_nr = it)})
+  })
+})

### 2020-09-08 R/impute.R
Combine ampute and impute steps for efficiency

@@ -1,27 +1,27 @@
 # Impute the data
 impute <-
   function(amp,
-           mids,
            m_mech,
            p_inc,
            it_nr,
-           it_total = 50,
-           n_imp = 5) {
+           it_total,
+           ...) {
     
     # initialize the mids object in the first iteration
-    if (it_nr == 1) {
+    if (is.null(mids)) {
       mids <<- mice(
-        amp[[m_mech]][[p_inc]],
+        amp,
         m = n_imp,
         method = "norm",
         maxit = 1,
         printFlag = FALSE
       )
-    }
-    
+    } else {
     # append to the mids object in any further iteration
-    if (it_nr > 1) {
-      mids <<- mice.mids(mids, printFlag = FALSE)
+      mids <<- mice.mids(
+        mids, 
+        maxit = 1, 
+        printFlag = FALSE)
     }
     
     # analyze and pool imputations
@@ -60,27 +60,11 @@ impute <-
     )
     
     # save in global environment
-    # if (it_nr == it_total) {
-    #   chain_means <<-
-    #     data.frame(
-    #       mech = rep(m_mech, it_total),
-    #       p = rep(p_inc, it_total),
-    #       it = 1:it_total,
-    #       chain.mean.Y = mids$chainMean["Y", ,],
-    #       chain.mean.X1 = mids$chainMean["X1", ,],
-    #       chain.mean.X2 = mids$chainMean["X2", ,]
-    #     )
-    #   
-    #   chain_vars <<-
-    #     data.frame(
-    #       mech = m_mech,
-    #       p = p_inc,
-    #       it = 1:it_total,
-    #       chain.var.Y = mids$chainVar["Y", ,],
-    #       chain.var.X1 = mids$chainVar["X1", ,],
-    #       chain.var.X2 = mids$chainVar["X2", ,]
-    #     )
-    # }
+    if (it_nr == it_total) {
+      chainmeans <<- c(chainmeans, list(mids$chainMean))
+      chainvars <<- c(chainvars, list(mids$chainVar))
+      mids <<- NULL
+    }
     return(out)
   }
 

### 2020-09-08 R/setup.R

@@ -24,8 +24,9 @@ setup <- function(seed) {
   
   # simulation parameters
   p_inc <<- c(0.05, 0.5, 0.95)
-  m_mech <<- c("MCAR", "MAR", "MCAR")
+  m_mech <<- c("MCAR", "MAR", "MNAR")
   
-  # object to store imputations in
-  mids <<- NULL
+  # objects to store imputations in
+  mids <<- chainmeans <<- chainvars <<- NULL
+  
 }

### 2020-09-08 RunSimulation.R

@@ -2,27 +2,54 @@
 
 # setup simulation environment
 source("R/setup.R")
+source("R/impute.R")
 setup(seed = 1111)
+it_total <- 4
+n_sim <- 3
+n_imp <- 5
 
+# create simulation function
+simulate <- function(dat, m_mech, p_inc, amp_pat, it_total, n_imp, ...){
 # ampute the data
-amp <- map(m_mech, function(mm) {
-  map(p_inc, function(pp) {
-    mice::ampute(
+# amp <- map(m_mech, function(mm) {
+#   map(p_inc, function(pp) {
+#     mice::ampute(
+#       data = dat,
+#       prop = pp,
+#       patterns = amp_pat,
+#       mech = mm
+#     )$amp
+#   }) %>% set_names(as.character(p_inc))
+# }) %>% set_names(as.character(m_mech))
+
+out <- map_dfr(m_mech, function(mm) {
+  map_dfr(p_inc, function(pp) {
+    # ampute the data
+    amp <- mice::ampute(
       data = dat,
       prop = pp,
       patterns = amp_pat,
-      mech = "MCAR"
+      mech = mm
     )$amp
-  }) %>% set_names(as.character(p_inc))
-}) %>% set_names(as.character(m_mech))
-
-results <- map_dfr(m_mech, function(mm) {
-  map_dfr(p_inc, function(pp) {
-    map_dfr(1:50, function(it) {
+    
+    map_dfr(1:it_total, function(it) {
+  # impute and extract results
   impute(amp,
-  mids,
   m_mech = mm,
   p_inc = pp,
-  it_nr = it)})
+  it_nr = it,
+  it_total,
+  chainmeans,
+  chainvars)})
   })
 })
+return(out)
+}
+
+# run simulation n_sim times
+res <-
+  replicate(n = n_sim,
+            expr = simulate(dat, m_mech, p_inc, amp_pat, it_total, chainmeans, chainvars),
+            simplify = FALSE)
+
+#a <- mids$chainMean %>% as.data.frame() %>% t() %>% as.data.frame() %>% cbind(it=1:it_total, m = rep(1:5, each = 5))

### 2020-09-08 R/preprocess.R
preprocess chain means and chain variances

@@ -1,0 +1,19 @@
+# preprocess theta values
+preprocess <- function(theta, ext) {
+  out <- map_dfr(m_mech, function(mm) {
+    map_dfr(p_inc, function(pp) {
+      chainmeans[[which(m_mech == mm) + which(p_inc == pp)]] %>%
+        as.data.frame() %>% t() %>% as.data.frame() %>%
+        setNames(paste0(ext, names(.))) %>%
+        cbind(
+          mech = mm,
+          p = pp,
+          it = 1:it_total,
+          m = rep(1:n_imp, each = it_total),
+          .
+        )
+    })
+  })
+  rownames(out) <- NULL
+  return(out)
+}

### 2020-09-08 RunSimulation.R

@@ -10,37 +10,26 @@ n_imp <- 5
 
 # create simulation function
 simulate <- function(dat, m_mech, p_inc, amp_pat, it_total, n_imp, ...){
-# ampute the data
-# amp <- map(m_mech, function(mm) {
-#   map(p_inc, function(pp) {
-#     mice::ampute(
-#       data = dat,
-#       prop = pp,
-#       patterns = amp_pat,
-#       mech = mm
-#     )$amp
-#   }) %>% set_names(as.character(p_inc))
-# }) %>% set_names(as.character(m_mech))
-
-out <- map_dfr(m_mech, function(mm) {
-  map_dfr(p_inc, function(pp) {
-    # ampute the data
+out <- 
+  # for each missingness mechanism...
+  map_dfr(m_mech, function(mm) {
+  # for each proportion of incomplete cases...
+    map_dfr(p_inc, function(pp) {
+    # ampute the complete data
     amp <- mice::ampute(
       data = dat,
       prop = pp,
       patterns = amp_pat,
       mech = mm
     )$amp
-    
+    # for each number of iterations...
     map_dfr(1:it_total, function(it) {
-  # impute and extract results
+  # impute the incomplete date and extract results
   impute(amp,
   m_mech = mm,
   p_inc = pp,
   it_nr = it,
-  it_total,
-  chainmeans,
-  chainvars)})
+  it_total)})
   })
 })
 return(out)
@@ -52,4 +41,14 @@ res <-
             expr = simulate(dat, m_mech, p_inc, amp_pat, it_total, chainmeans, chainvars),
             simplify = FALSE)
 
-#a <- mids$chainMean %>% as.data.frame() %>% t() %>% as.data.frame() %>% cbind(it=1:it_total, m = rep(1:5, each = 5))
+# extract chain means and chain variances
+mus <- preprocess(theta = chainmeans, ext = "mu.")
+sigmas <- preprocess(theta = chainvars, ext = "sigma.")
+
+# apply convergence diagnostics
+# [some code here]
+
+# summarize results
+tab <- map_df(res, ~ {
+  as.data.frame(.)
+}) %>% aggregate(. ~ it + p + mech, data = ., mean)

### 2020-09-08 R/ComputeAC.R
Split convergence diagnostics and other outcomes

@@ -1,0 +1,43 @@
+# extract convergence diagnostic autocorrelation from iteration object (e.g., chain means)
+# requires packages 'purrr' and 'magrittr'/'dplyr', and the functions in 'SupplementAC.R'
+
+# load supplementary file
+source('Thesis/Functions/SupplementAC.R')
+
+# function for lag 1 AC for one or more imputation chains
+ac_adapted <- function(x, ac_function = "ac") {
+  # input: object with theta values (rows are iterations, columns are imputations),
+  # and preferred way of calculating ac values
+  # output: autocorrelations at each iteration in x, computed per imputation chain
+  # and with max and mean across imputations
+  
+  # set function to apply
+  if (ac_function == "acf") {
+    function_to_apply <- acf_lag1
+    min_it <- 2
+  } else {
+    function_to_apply <- ac_lag1
+    min_it <- 3
+  }
+  
+  # make input ready for mapping
+  if (is.numeric(x)) {
+    x <- as.data.frame(x)
+  }
+  t <- dim(x)[1] #nr of iterations
+  M <- dim(x)[2] #nr of imputations
+  
+  # compute ac for each iteration (by definition only possible for t > 2)
+  if (t < min_it) {
+    ac <-
+      matrix(NA, t, M + 2) %>% as.data.frame %>% set_names(paste0(ac_function, ".", c("mean", "max", names(x))))
+  } else {
+    ac <- map_dfr(min_it:t, function(it) {
+      # compute ac
+      x[1:it, ] %>% function_to_apply() %>% t() %>% data.frame(mean(.), max(.), .) %>% set_names(paste0(ac_function, ".", c("mean", "max", 1:M)))
+    }) %>% rbind(matrix(NA, min_it-1, M+2, dimnames = list(NULL, names(.))), .) %>% cbind(iteration = 1:t, .)
+  }
+  
+  # output
+  return(ac)
+}

### 2020-09-08 R/ComputeDiagnostics.R

@@ -1,0 +1,16 @@
+# function to compute Rhat conform Vehtari et al (2019) and calculate AC manually
+# requires 'dplyr', and the functions in 'ComputeAC.R', and 'ComputeRhat.R'
+
+# load functions
+source('Thesis/Functions/ComputeRhat.R')
+source('Thesis/Functions/ComputeAC.R')
+
+# combine the two diagnostics
+convergence <- function(x, include_acf = FALSE) {
+  # input: object with imputation chain(s), e.g. chain mean, choice of calculating default AC
+  # output: AC values at each iteration, computed manually, if necessary also using stats:acf() 
+  out <-  left_join(rhat_adapted(x), ac_adapted(x), by = "iteration") 
+  if(include_acf){ out <- left_join(out, ac_adapted(x, "acf"), by = "iteration")
+  }
+  return(out)
+}

### 2020-09-08 R/ComputeRhat.R

@@ -1,0 +1,38 @@
+# extract convergence diagnostic Rhat from iteration object (e.g., chain means)
+# requires packages 'purrr' and 'magrittr'/'dplyr', and the functions in SupplementRhat.R
+
+# load function
+source('Thesis/Functions/SupplementRhat.R')
+
+# function for Rhat for two or more imputation chains
+rhat_adapted <- function(x) {
+  # input: object with theta values (rows are iterations, columns are imputations)
+  # output: convergence diagnostic Rhat across imputations
+  
+  # make input ready for mapping
+  if (is.data.frame(x)) {
+    x <- as.matrix(x)
+  }
+  
+  # quit function if there are not enough iterations
+  t <- dim(x)[1]
+  if (is.null(t)) {
+    r.hat <- data.frame(iteration = 1,
+                        max.r.hat = NA,
+                        r.hat = NA)
+  }
+  else {
+    r.hat <- map_dfr(2:t, function(it) {
+      # compute original r hat conform Gelman and Rubin (1992) 
+      rhat_original <- x[1:it,] %>% get.rhat()
+      # compute r hat in all ways described by Vehtari et al. (2019)
+      rhat_bulk <-
+        x[1:it,] %>% split_chains() %>% z_scale() %>% get.rhat()
+      rhat_tail <-
+        x[1:it,] %>% fold_sims() %>% split_chains() %>% z_scale() %>% get.rhat()
+      max(rhat_bulk, rhat_tail) %>% data.frame(r.hat.max = ., r.hat = rhat_original) #%>% set_names("r.hat")
+    }) %>% rbind(NA, .) %>% cbind(iteration = 1:t, .)
+  }
+  # output
+  return(r.hat)
+}

### 2020-09-08 R/impute.R

@@ -36,7 +36,7 @@ impute <-
     ci_lo <- est - qt(.975, df = n_imp - 1) * SE
     ci_up <- est + qt(.975, df = n_imp - 1) * SE
     
-    # add convergence parameters
+    # get convergence parameters
     mild <- mids %>% mice::complete("all")
     qhat <-
       map_dbl(mild, ~ {
@@ -54,11 +54,24 @@ impute <-
       it = it_nr,
       est = est,
       CIW = ci_up - ci_lo,
-      cov = ci_lo < Q & Q < ci_up,
-      qhat = t(qhat),
-      lambda = t(lambda)
+      cov = ci_lo < Q & Q < ci_up#,
+      #qhat = t(qhat),
+      #lambda = t(lambda)
     )
-    
+    qhats <<- rbind(qhats, data.frame(
+      mech = m_mech,
+      p = p_inc,
+      it = it_nr,
+      m = 1:5,
+      qhat = qhat
+    ))
+    lambdas <<- rbind(lambdas, data.frame(
+      mech = m_mech,
+      p = p_inc,
+      it = it_nr,
+      m = 1:n_imp,
+      lambda = lambda
+    ))
     # save in global environment
     if (it_nr == it_total) {
       chainmeans <<- c(chainmeans, list(mids$chainMean))

### 2020-09-08 R/preprocess.R

@@ -1,8 +1,8 @@
 # preprocess theta values
-preprocess <- function(theta, ext) {
+preprocess <- function(theta, ext, ...) {
   out <- map_dfr(m_mech, function(mm) {
     map_dfr(p_inc, function(pp) {
-      chainmeans[[which(m_mech == mm) + which(p_inc == pp)]] %>%
+      theta[[which(m_mech == mm) + which(p_inc == pp)]] %>%
         as.data.frame() %>% t() %>% as.data.frame() %>%
         setNames(paste0(ext, names(.))) %>%
         cbind(

### 2020-09-08 R/setup.R

@@ -27,6 +27,6 @@ setup <- function(seed) {
   m_mech <<- c("MCAR", "MAR", "MNAR")
   
   # objects to store imputations in
-  mids <<- chainmeans <<- chainvars <<- NULL
+  mids <<- chainmeans <<- chainvars <<- qhats <<- lambdas <<- NULL
   
 }

### 2020-09-08 RunSimulation.R

@@ -3,9 +3,10 @@
 # setup simulation environment
 source("R/setup.R")
 source("R/impute.R")
+source("R/preprocess.R")
 setup(seed = 1111)
 it_total <- 4
-n_sim <- 3
+n_sim <- 2
 n_imp <- 5
 
 # create simulation function
@@ -29,7 +30,9 @@ out <-
   m_mech = mm,
   p_inc = pp,
   it_nr = it,
-  it_total)})
+  it_total,
+  chainmeans,
+  chainvars)})
   })
 })
 return(out)
@@ -41,9 +44,11 @@ res <-
             expr = simulate(dat, m_mech, p_inc, amp_pat, it_total, chainmeans, chainvars),
             simplify = FALSE)
 
-# extract chain means and chain variances
+# preprocess theta values
 mus <- preprocess(theta = chainmeans, ext = "mu.")
 sigmas <- preprocess(theta = chainvars, ext = "sigma.")
+qhats <- cbind(sim = rep(1:n_sim, each = n_imp*it_total*length(m_mech)*length(p_inc)), qhats)
+lambdas <- cbind(sim = rep(1:n_sim, each = n_imp*it_total*length(m_mech)*length(p_inc)), lambdas)
 
 # apply convergence diagnostics
 # [some code here]
@@ -51,4 +56,4 @@ sigmas <- preprocess(theta = chainvars, ext = "sigma.")
 # summarize results
 tab <- map_df(res, ~ {
   as.data.frame(.)
-}) %>% aggregate(. ~ it + p + mech, data = ., mean)
+}) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)})

### 2020-09-11 Manuscript.Rmd
Add convergence diagnostics to outcomes

@@
 
-- Missing data has to be accomodated, usually filled in
+- Missing data has to be accommodated, usually filled in
 
@@
 
-- In this paper we explore diagnostic methods for MICE, and show which parameter has the best performance
+- In this paper we explore diagnostic methods for MICE, and show which parameter has the best performance --> most informative about convergence
 
@@
 
-- outcomes: estimated regression coefficient, CIW, coverage, R2
+- outcomes: estimated regression coefficient, CIW, coverage, R2 --> multivariate is key, because univariate convergence does not guarantee multivariate convergence
 

### 2020-09-11 R/ComputeAC.R

@@ -2,7 +2,7 @@
 # requires packages 'purrr' and 'magrittr'/'dplyr', and the functions in 'SupplementAC.R'
 
 # load supplementary file
-source('Thesis/Functions/SupplementAC.R')
+source('R/SupplementAC.R')
 
 # function for lag 1 AC for one or more imputation chains
 ac_adapted <- function(x, ac_function = "ac") {

### 2020-09-11 R/ComputeDiagnostics.R

@@ -2,8 +2,8 @@
 # requires 'dplyr', and the functions in 'ComputeAC.R', and 'ComputeRhat.R'
 
 # load functions
-source('Thesis/Functions/ComputeRhat.R')
-source('Thesis/Functions/ComputeAC.R')
+source('R/ComputeRhat.R')
+source('R/ComputeAC.R')
 
 # combine the two diagnostics
 convergence <- function(x, include_acf = FALSE) {

### 2020-09-11 R/ComputeRhat.R

@@ -2,7 +2,7 @@
 # requires packages 'purrr' and 'magrittr'/'dplyr', and the functions in SupplementRhat.R
 
 # load function
-source('Thesis/Functions/SupplementRhat.R')
+source('R/SupplementRhat.R')
 
 # function for Rhat for two or more imputation chains
 rhat_adapted <- function(x) {

### 2020-09-11 R/preprocess.R

@@ -1,11 +1,13 @@
 # preprocess theta values
 preprocess <- function(theta, ext, ...) {
-  out <- map_dfr(m_mech, function(mm) {
-    map_dfr(p_inc, function(pp) {
-      theta[[which(m_mech == mm) + which(p_inc == pp)]] %>%
+  out <- map_dfr(1:n_sim, function(i){
+    map_dfr(m_mech, function(mm) {
+      map_dfr(p_inc, function(pp) {
+      theta[[i * which(m_mech == mm) * which(p_inc == pp)]] %>%
         as.data.frame() %>% t() %>% as.data.frame() %>%
         setNames(paste0(ext, names(.))) %>%
         cbind(
+          sim = i,
           mech = mm,
           p = pp,
           it = 1:it_total,
@@ -13,6 +15,7 @@ preprocess <- function(theta, ext, ...) {
           .
         )
     })
+  })
   })
   rownames(out) <- NULL
   return(out)

### 2020-09-11 RunSimulation.R

@@ -49,11 +49,24 @@ mus <- preprocess(theta = chainmeans, ext = "mu.")
 sigmas <- preprocess(theta = chainvars, ext = "sigma.")
 qhats <- cbind(sim = rep(1:n_sim, each = n_imp*it_total*length(m_mech)*length(p_inc)), qhats)
 lambdas <- cbind(sim = rep(1:n_sim, each = n_imp*it_total*length(m_mech)*length(p_inc)), lambdas)
+thetas <- mus %>% 
+  dplyr::full_join(sigmas) %>% 
+  dplyr::full_join(qhats) %>% 
+  dplyr::full_join(lambdas) 
 
-# apply convergence diagnostics
-# [some code here]
+# apply convergence diagnostics to each theta
+# thetas %>% 
+#   select(sim, mech, p, m, it, "lambda") %>% 
+#   filter(mech == "MCAR", p == 0.5, sim == 1) %>%  
+#   pivot_wider(names_from = m, values_from = "lambda") %>% 
+#   .[,-c(1:4)] %>% 
+#   convergence()
 
 # summarize results
 tab <- map_df(res, ~ {
   as.data.frame(.)
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)})
+
+save(thetas, file = "Data/parameters.Rdata")
+save(res, file = "Data/outcomes.Rdata")
+

### 2020-09-11 Manuscript.Rmd
Simplified simulation

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(tidyverse)
 ```
@@
 - diagnostics: rhat, ac
+
+```{r echo=FALSE}
+load("~/MissingThePoint/Data/results.Rdata")
+results %>% 
+  ggplot() + 
+  geom_hline(yintercept = 0.3334) +
+  geom_point(aes(x=it, y=est, color = mech, shape = as.factor(p))) + 
+  geom_line(aes(x=it, y=est, color = mech, linetype = as.factor(p))) + 
+  theme_classic()
+```

### 2020-09-11 R/impute.R

@@ -25,8 +25,9 @@ impute <-
     }
     
     # analyze and pool imputations
-    mipo <- mids %>%
-      mice::lm.mids(Y ~ ., .) %>%
+    mira <- mids %>%
+      mice::lm.mids(Y ~ ., .) 
+    mipo <- mira %>%
       mice::pool() %>%
       .$pooled
     
@@ -35,6 +36,7 @@ impute <-
     SE <- sqrt(mipo$b + (mipo$b / n_imp)) %>% .[2]
     ci_lo <- est - qt(.975, df = n_imp - 1) * SE
     ci_up <- est + qt(.975, df = n_imp - 1) * SE
+    r_sq <- pool.r.squared(mira) %>% .[1]
     
     # get convergence parameters
     mild <- mids %>% mice::complete("all")
@@ -54,7 +56,8 @@ impute <-
       it = it_nr,
       est = est,
       CIW = ci_up - ci_lo,
-      cov = ci_lo < Q & Q < ci_up#,
+      cov = ci_lo < Q & Q < ci_up,
+      rsq = r_sq
       #qhat = t(qhat),
       #lambda = t(lambda)
     )

### 2020-09-11 RunSimulation.R

@@ -5,8 +5,8 @@ source("R/setup.R")
 source("R/impute.R")
 source("R/preprocess.R")
 setup(seed = 1111)
-it_total <- 4
-n_sim <- 2
+it_total <- 50
+n_sim <- 100
 n_imp <- 5
 
 # create simulation function
@@ -39,7 +39,7 @@ return(out)
 }
 
 # run simulation n_sim times
-res <-
+outcomes <-
   replicate(n = n_sim,
             expr = simulate(dat, m_mech, p_inc, amp_pat, it_total, chainmeans, chainvars),
             simplify = FALSE)
@@ -63,10 +63,10 @@ thetas <- mus %>%
 #   convergence()
 
 # summarize results
-tab <- map_df(res, ~ {
+results <- map_df(outcomes, ~ {
   as.data.frame(.)
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)})
 
 save(thetas, file = "Data/parameters.Rdata")
-save(res, file = "Data/outcomes.Rdata")
-
+save(outcomes, file = "Data/outcomes.Rdata")
+save(results, file = "Data/results.Rdata")

### 2020-09-14 Manuscript.Rmd
Plot outcomes simulation

--- 2020-09-11 Manuscript.Rmd
@@
 - diagnostics: rhat, ac
+
+## Some results
+
+```{r echo=FALSE}
+load("~/MissingThePoint/Data/results.Rdata")
+results %>%
+  ggplot() +
+  geom_hline(yintercept = 0.3334) +
+  geom_point(aes(x=it, y=est, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=est, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+# results %>% 
+#   ggplot() + 
+#   geom_hline(yintercept = 0.95) +
+#   geom_point(aes(x=it, y=cov, color = mech, shape = as.factor(p))) + 
+#   geom_line(aes(x=it, y=cov, color = mech, linetype = as.factor(p))) + 
+#   theme_classic()
+
+results %>%
+  ggplot() +
+  #geom_hline(yintercept = 0.3334) +
+  geom_point(aes(x=it, y=CIW, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=CIW, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+results %>%
+  ggplot() +
+  geom_hline(yintercept = 0.2090) +
+  geom_point(aes(x=it, y=rsq, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=rsq, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```

### 2020-09-14 R/setup.R

@@ -15,6 +15,9 @@ setup <- function(seed) {
     broom::tidy() %>%
     .[["estimate"]] %>%
     .[2]
+  r2 <- lm(Y ~ X1 + X2, dat) %>% 
+    summary() %>% 
+    .[["r.squared"]]
   
   # create patterns to ampute the data with multivariate missingness
   amp_pat <<-

### 2020-09-16 Manuscript.Rmd
Compute convergence diagnostics and plot results

--- 2020-09-11 Manuscript.Rmd
@@
 - diagnostics: rhat, ac
+
+## Some results
+
+```{r echo=FALSE}
+load("~/MissingThePoint/Data/results.Rdata")
+results %>%
+  ggplot() +
+  geom_hline(yintercept = 0.3334) +
+  geom_point(aes(x=it, y=est, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=est, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+
+Bias in estimate depends mostly on the amount of missingness (less info to estimate the relation??), but overall MNAR performs the worst, which is stable after it=5.
+
+```{r}
+results %>%
+  ggplot() +
+  geom_hline(yintercept = 0.95) +
+  geom_point(aes(x=it, y=cov, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=cov, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+
+Cov is more or less fine, except for MNAR with only 0.05% incomplete cases --> due to bias in the estimate in combination with very narrow CI (see CIW!). However, all stable after it=3??
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=CIW, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=CIW, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+
+CIW clearly determined by the proportion of incomplete cases. Worst for p=0.95, then stable after it 7.
+
+```{r}
+results %>%
+  ggplot() +
+  geom_hline(yintercept = 0.2090) +
+  geom_point(aes(x=it, y=rsq, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=rsq, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+
+Magnitude of the bias in R^2 perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for p=0.95, MNAR, but stable from it=5.
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=ac.max.lambda, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=ac.max.lambda, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+```
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=r.hat.max.lambda, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=r.hat.max.lambda, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```

### 2020-09-16 R/ComputeDiagnostics.R

@@ -12,5 +12,6 @@ convergence <- function(x, include_acf = FALSE) {
   out <-  left_join(rhat_adapted(x), ac_adapted(x), by = "iteration") 
   if(include_acf){ out <- left_join(out, ac_adapted(x, "acf"), by = "iteration")
   }
+  out <- out %>% dplyr::rename(it = iteration)
   return(out)
 }

### 2020-09-16 RunSimulation.R

@@ -4,10 +4,12 @@
 source("R/setup.R")
 source("R/impute.R")
 source("R/preprocess.R")
+source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
 it_total <- 50
 n_sim <- 100
 n_imp <- 5
+thetas <- c("mu.Y", "mu.X1", "mu.X2", "sigma.Y", "sigma.X1", "sigma.X2", "qhat", "lambda")
 
 # create simulation function
 simulate <- function(dat, m_mech, p_inc, amp_pat, it_total, n_imp, ...){
@@ -49,12 +51,34 @@ mus <- preprocess(theta = chainmeans, ext = "mu.")
 sigmas <- preprocess(theta = chainvars, ext = "sigma.")
 qhats <- cbind(sim = rep(1:n_sim, each = n_imp*it_total*length(m_mech)*length(p_inc)), qhats)
 lambdas <- cbind(sim = rep(1:n_sim, each = n_imp*it_total*length(m_mech)*length(p_inc)), lambdas)
-thetas <- mus %>% 
+parameters <- mus %>% 
   dplyr::full_join(sigmas) %>% 
   dplyr::full_join(qhats) %>% 
   dplyr::full_join(lambdas) 
 
 # apply convergence diagnostics to each theta
+convergence_diagnostics <- purrr::map_dfr(1:n_sim, function(ss) {
+  purrr::map_dfr(m_mech, function(mm) {
+    purrr::map_dfr(p_inc, function(pp) {
+      purrr::map_dfc(thetas, function(tt) {
+        one_theta <- parameters %>%
+          filter(sim == ss, mech == mm, p == pp) %>%
+          select(it, m, tt) %>%
+          arrange(it)
+        matrix(one_theta[[tt]], ncol = n_imp, byrow = T) %>%
+          convergence() %>%
+          select(r.hat.max, ac.max) %>%
+          setNames(., paste0(names(.), ".", tt))
+      }) %>% cbind(
+        sim = ss,
+        mech = mm,
+        p = pp,
+        it = 1:it_total,
+        .
+      )
+    })
+  })
+})
 # thetas %>% 
 #   select(sim, mech, p, m, it, "lambda") %>% 
 #   filter(mech == "MCAR", p == 0.5, sim == 1) %>%  
@@ -63,10 +87,15 @@ thetas <- mus %>%
 #   convergence()
 
 # summarize results
[…]

### 2020-09-17 Manuscript.Rmd
Run sim with different CI computation to check why coverages are so off (n_sim = 100, maxit = 5)

@@
 library(tidyverse)
+load("~/MissingThePoint/Data/results.Rdata")
+
 ```
@@
 ```{r echo=FALSE}
-load("~/MissingThePoint/Data/results.Rdata")
 results %>%
@@
 Bias in estimate depends mostly on the amount of missingness (less info to estimate the relation??), but overall MNAR performs the worst, which is stable after it=5.
+
+Update with different CI computation, see https://stefvanbuuren.name/fimd/sec-evaluation.html#sec:quantifyingbias
 
@@
   ggplot() +
+  geom_point(aes(x=it, y=ac.max.mu.Y, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=ac.max.mu.Y, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+```
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=r.hat.max.mu.Y, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=r.hat.max.mu.Y, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=ac.max.sigma.Y, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=ac.max.sigma.Y, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+```
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=r.hat.max.sigma.Y, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=r.hat.max.sigma.Y, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=ac.max.qhat, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=ac.max.qhat, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+```
+
+```{r}
+results %>%
+  ggplot() +
+  geom_point(aes(x=it, y=r.hat.max.qhat, color = mech, shape = as.factor(p))) +
+  geom_line(aes(x=it, y=r.hat.max.qhat, color = mech, linetype = as.factor(p))) +
+  theme_classic()
+
+```
+
+```{r}
+results %>%
+  ggplot() +
   geom_point(aes(x=it, y=ac.max.lambda, color = mech, shape = as.factor(p))) +
@@
 ```
+
+## Discussion
+
+- Estimates etc do not improve after it = 7
+
+- Convergence diagnostics keep improving, but are all stable after 20-30 it
+
+- Univariate thetas do not differentiate between missingness mechanisms or proportions of incomplete cases
+
+- Determining non-stationarity with lambda is more difficult than with qhat :(

### 2020-09-17 R/impute.R

@@ -29,13 +29,13 @@ impute <-
       mice::lm.mids(Y ~ ., .) 
     mipo <- mira %>%
       mice::pool() %>%
-      .$pooled
+      summary(., conf.int = TRUE)
     
     # extract statistics
-    est <- mipo$estimate[2]
-    SE <- sqrt(mipo$b + (mipo$b / n_imp)) %>% .[2]
-    ci_lo <- est - qt(.975, df = n_imp - 1) * SE
-    ci_up <- est + qt(.975, df = n_imp - 1) * SE
+    est <- mipo[["estimate"]][2]
+    # SE <- sqrt(mipo$b + (mipo$b / n_imp)) %>% .[2]
+    ci_lo <- mipo[["2.5 %"]][2] # est - qt(.975, df = n_imp - 1) * SE
+    ci_up <- mipo[["97.5 %"]][2] # est + qt(.975, df = n_imp - 1) * SE
     r_sq <- pool.r.squared(mira) %>% .[1]
     
     # get convergence parameters

### 2020-09-17 RunSimulation.R

@@ -6,7 +6,7 @@ source("R/impute.R")
 source("R/preprocess.R")
 source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
-it_total <- 50
+it_total <- 5
 n_sim <- 100
 n_imp <- 5
 thetas <- c("mu.Y", "mu.X1", "mu.X2", "sigma.Y", "sigma.X1", "sigma.X2", "qhat", "lambda")
@@ -96,6 +96,10 @@ results <- map_df(outcomes, ~ {
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)}) %>% 
   full_join(., conv_results)
 
-save(parameters, file = "Data/parameters.Rdata")
-save(outcomes, file = "Data/outcomes.Rdata")
-save(results, file = "Data/results.Rdata")
+save(parameters, file = "Data/parameters_standard_ci.Rdata")
+save(outcomes, file = "Data/outcomes_standard_ci.Rdata")
+save(results, file = "Data/results_standard_ci.Rdata")
+
+# save(parameters, file = "Data/parameters.Rdata")
+# save(outcomes, file = "Data/outcomes.Rdata")
+# save(results, file = "Data/results.Rdata")

### 2020-09-17 RunSimulation.R
Same but 100 sims

@@ -8,7 +8,7 @@ source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
 it_total <- 5
 n_sim <- 100
-n_imp <- 5
+n_imp <- 50
 thetas <- c("mu.Y", "mu.X1", "mu.X2", "sigma.Y", "sigma.X1", "sigma.X2", "qhat", "lambda")
 
 # create simulation function

### 2020-09-18 Manuscript.Rmd
Include data generation in each simulation rep, to suit the regular CI computation

@@
 library(tidyverse)
-load("~/MissingThePoint/Data/results.Rdata")
+load("~/MissingThePoint/Data/results_standard_ci.Rdata")
 
@@
   ggplot() +
-  geom_hline(yintercept = 0.3334) +
+  #geom_hline(yintercept = 0.3334) +
   geom_point(aes(x=it, y=est, color = mech, shape = as.factor(p))) +
@@
   ggplot() +
-  geom_hline(yintercept = 0.95) +
+  #geom_hline(yintercept = 0.95) +
   geom_point(aes(x=it, y=cov, color = mech, shape = as.factor(p))) +
@@
   ggplot() +
-  geom_hline(yintercept = 0.2090) +
+  #geom_hline(yintercept = 0.2090) +
   geom_point(aes(x=it, y=rsq, color = mech, shape = as.factor(p))) +
@@
 ```
+
 ```{r}

### 2020-09-18 R/generate.R

@@ -1,0 +1,15 @@
+# data generating
+
+generate <- function(N, vcov){
+dat <<- mvtnorm::rmvnorm(N, sigma = vcov) %>%
+  as.data.frame() %>%
+  setNames(c("Y", "X1", "X2")) %>%
+  mutate(Y = Y + rnorm(N, sd = 1))
+Q <<- lm(Y ~ X1 + X2, dat) %>%
+  broom::tidy() %>%
+  .[["estimate"]] %>%
+  .[2]
+r2 <<- lm(Y ~ X1 + X2, dat) %>% 
+  summary() %>% 
+  .[["r.squared"]]
+}

### 2020-09-18 R/setup.R

@@ -5,29 +5,21 @@ setup <- function(seed) {
   library(tidyverse)
   set.seed(seed)
   
-  # create complete data
-  vcov <- matrix(c(1, .5, .5, .5, 1, .5, .5, .5, 1), ncol = 3)
-  dat <<- mvtnorm::rmvnorm(1000, sigma = vcov) %>%
-    as.data.frame() %>%
-    setNames(c("Y", "X1", "X2")) %>%
-    mutate(Y = Y + rnorm(1000, sd = 1))
-  Q <<- lm(Y ~ X1 + X2, dat) %>%
-    broom::tidy() %>%
-    .[["estimate"]] %>%
-    .[2]
-  r2 <- lm(Y ~ X1 + X2, dat) %>% 
-    summary() %>% 
-    .[["r.squared"]]
+  # complete data variance-covariance matrix
+  varcov <<- matrix(c(1, .5, .5, .5, 1, .5, .5, .5, 1), ncol = 3)
   
-  # create patterns to ampute the data with multivariate missingness
+  # patterns to ampute the data with multivariate missingness
   amp_pat <<-
     expand.grid(c(0, 1), c(0, 1), c(0, 1)) %>% #define all possible combinations of univariate and multivariate missingness
     .[c(-1, -8),]  %>% #remove the completely (un)observed cases
-    setNames(names(ampute(dat)$patterns)) #obtain correct names of patterns
+    setNames(c("Y", "X1", "X2")) #give correct names 
   
   # simulation parameters
-  p_inc <<- c(0.05, 0.5, 0.95)
+  n_cases <<- 1000
+  n_imp <<- 5
   m_mech <<- c("MCAR", "MAR", "MNAR")
+  p_inc <<- c(0.05, 0.5, 0.95)
+  thetas <<- c("mu.Y", "mu.X1", "mu.X2", "sigma.Y", "sigma.X1", "sigma.X2", "qhat", "lambda")
   
   # objects to store imputations in
   mids <<- chainmeans <<- chainvars <<- qhats <<- lambdas <<- NULL

### 2020-09-18 RunSimulation.R

@@ -2,17 +2,17 @@
 
 # setup simulation environment
 source("R/setup.R")
+source("R/generate.R")
 source("R/impute.R")
 source("R/preprocess.R")
 source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
-it_total <- 5
+it_total <- 50
 n_sim <- 100
-n_imp <- 50
-thetas <- c("mu.Y", "mu.X1", "mu.X2", "sigma.Y", "sigma.X1", "sigma.X2", "qhat", "lambda")
 
 # create simulation function
 simulate <- function(dat, m_mech, p_inc, amp_pat, it_total, n_imp, ...){
+  generate(N=n_cases, vcov = varcov)
 out <- 
   # for each missingness mechanism...
   map_dfr(m_mech, function(mm) {
@@ -96,9 +96,9 @@ results <- map_df(outcomes, ~ {
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)}) %>% 
   full_join(., conv_results)
 
-save(parameters, file = "Data/parameters_standard_ci.Rdata")
-save(outcomes, file = "Data/outcomes_standard_ci.Rdata")
-save(results, file = "Data/results_standard_ci.Rdata")
+save(parameters, file = "Data/parameters_DGM.Rdata")
+save(outcomes, file = "Data/outcomes_DGM.Rdata")
+save(results, file = "Data/results_DGM.Rdata")
 
 # save(parameters, file = "Data/parameters.Rdata")
 # save(outcomes, file = "Data/outcomes.Rdata")

### 2020-09-18 Manuscript.Rmd
Edit abstract

@@
 
-# Missing The Point
+# Missing The Point: Non-Convergence in Iterative Imputation Algorithms
+
+## Abstract
+
+Iterative imputation is a popular tool to accommodate the ubiquitous problem of missing data.
+While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
+Our study provides insight into identifying non-convergence in iterative imputation algorithms, since there is no consensus on how to evaluate the convergence properties currently.
+We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
+We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
 
 ## Intro
+
+Missing data poses a ubiquitous threat against unbiased, confidence-valid statistical inferences.
+A popular tool to accommodate missing data.
+With techniques like multiple imputation (Rubin, 1987), missing data points are imputed (i.e., 'filled in')
 

### 2020-09-21 Manuscript.Rmd
Simulation results with data generation in each run (n_sim = 100)
rerun sim after feedback mice meeting that coverages are off

@@
 library(tidyverse)
-load("~/MissingThePoint/Data/results_standard_ci.Rdata")
+load("~/MissingThePoint/Data/results.Rdata")
 
@@
 Missing data poses a ubiquitous threat against unbiased, confidence-valid statistical inferences.
-A popular tool to accommodate missing data.
+A popular tool to accommodate missing data is iterative imputation.
 With techniques like multiple imputation (Rubin, 1987), missing data points are imputed (i.e., 'filled in')

### 2020-09-21 RunSimulation.R

@@ -7,17 +7,17 @@ source("R/impute.R")
 source("R/preprocess.R")
 source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
-it_total <- 50
+it_total <- 5
 n_sim <- 100
 
 # create simulation function
 simulate <- function(dat, m_mech, p_inc, amp_pat, it_total, n_imp, ...){
-  generate(N=n_cases, vcov = varcov)
 out <- 
   # for each missingness mechanism...
   map_dfr(m_mech, function(mm) {
   # for each proportion of incomplete cases...
     map_dfr(p_inc, function(pp) {
+      generate(N=n_cases, vcov = varcov)
     # ampute the complete data
     amp <- mice::ampute(
       data = dat,
@@ -96,9 +96,9 @@ results <- map_df(outcomes, ~ {
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)}) %>% 
   full_join(., conv_results)
 
-save(parameters, file = "Data/parameters_DGM.Rdata")
-save(outcomes, file = "Data/outcomes_DGM.Rdata")
-save(results, file = "Data/results_DGM.Rdata")
+save(parameters, file = "Data/parameters_dat.Rdata")
+save(outcomes, file = "Data/outcomes_dat.Rdata")
+save(results, file = "Data/results_dat.Rdata")
 
 # save(parameters, file = "Data/parameters.Rdata")
 # save(outcomes, file = "Data/outcomes.Rdata")

### 2020-09-21 RunSimulation.R
Data generation in each simulation run, 100 sims, 50 it

@@ -7,7 +7,7 @@ source("R/impute.R")
 source("R/preprocess.R")
 source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
-it_total <- 5
+it_total <- 50
 n_sim <- 100
 
 # create simulation function
@@ -96,9 +96,9 @@ results <- map_df(outcomes, ~ {
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)}) %>% 
   full_join(., conv_results)
 
-save(parameters, file = "Data/parameters_dat.Rdata")
-save(outcomes, file = "Data/outcomes_dat.Rdata")
-save(results, file = "Data/results_dat.Rdata")
+save(parameters, file = "Data/parameters_cov.Rdata")
+save(outcomes, file = "Data/outcomes_cov.Rdata")
+save(results, file = "Data/results_cov.Rdata")
 
 # save(parameters, file = "Data/parameters.Rdata")
 # save(outcomes, file = "Data/outcomes.Rdata")

### 2020-09-21 R/generate.R
Change data generating mech to investigate coverage rates

@@ -1,15 +1,15 @@
 # data generating
 
 generate <- function(N, vcov){
-dat <<- mvtnorm::rmvnorm(N, sigma = vcov) %>%
+dat <<- mvtnorm::rmvnorm(n = N, sigma = vcov) %>%
   as.data.frame() %>%
-  setNames(c("Y", "X1", "X2")) %>%
-  mutate(Y = Y + rnorm(N, sd = 1))
-Q <<- lm(Y ~ X1 + X2, dat) %>%
+  setNames(c("Y", "X1", "X2")) #%>%
+  #mutate(Y = Y + rnorm(N))
+Q <<- lm(Y ~ ., dat) %>%
   broom::tidy() %>%
   .[["estimate"]] %>%
   .[2]
-r2 <<- lm(Y ~ X1 + X2, dat) %>% 
+r2 <<- lm(Y ~ ., dat) %>% 
   summary() %>% 
   .[["r.squared"]]
 }

### 2020-09-21 RunSimulation.R

@@ -11,7 +11,7 @@ it_total <- 50
 n_sim <- 100
 
 # create simulation function
-simulate <- function(dat, m_mech, p_inc, amp_pat, it_total, n_imp, ...){
+simulate <- function(m_mech, p_inc, amp_pat, it_total, n_imp, ...){
 out <- 
   # for each missingness mechanism...
   map_dfr(m_mech, function(mm) {
@@ -43,7 +43,7 @@ return(out)
 # run simulation n_sim times
 outcomes <-
   replicate(n = n_sim,
-            expr = simulate(dat, m_mech, p_inc, amp_pat, it_total, chainmeans, chainvars),
+            expr = simulate(m_mech, p_inc, amp_pat, it_total, chainmeans, chainvars),
             simplify = FALSE)
 
 # preprocess theta values

### 2020-09-22 Manuscript.Rmd
Data generation in each simulation run, 100 sims, 50 it
Set scientific estimate to 1 to compute coverage rates like they should

@@
 library(tidyverse)
-load("~/MissingThePoint/Data/results.Rdata")
+load("~/MissingThePoint/Data/results_test.Rdata")
 
@@
 ```{r}
-results %>%
+results %>% #filter(it>6) %>% 
   ggplot() +
-  #geom_hline(yintercept = 0.95) +
+  geom_hline(yintercept = 0.95, size = 2, color = "grey") +
   geom_point(aes(x=it, y=cov, color = mech, shape = as.factor(p))) +
@@
 results %>% 
+  # filter(it>6) %>% 
   ggplot() +
-  #geom_hline(yintercept = 0.2090) +
   geom_point(aes(x=it, y=rsq, color = mech, shape = as.factor(p))) +
@@
 - Determining non-stationarity with lambda is more difficult than with qhat :(
+
+## Check if heuristic for AC works
+```{r}
+load("~/MissingThePoint/Data/convergence_diagnostics_test.Rdata")
+convergence_diagnostics %>% filter(sim == 1, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line()
+convergence_diagnostics %>% filter(sim == 1, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line()
+```

### 2020-09-22 R/generate.R

@@ -3,12 +3,12 @@
 generate <- function(N, vcov){
 dat <<- mvtnorm::rmvnorm(n = N, sigma = vcov) %>%
   as.data.frame() %>%
-  setNames(c("Y", "X1", "X2")) #%>%
-  #mutate(Y = Y + rnorm(N))
-Q <<- lm(Y ~ ., dat) %>%
-  broom::tidy() %>%
-  .[["estimate"]] %>%
-  .[2]
+  setNames(c("Y", "X1", "X2")) %>%
+  mutate(Y = X1 - X2 + rnorm(N))
+# Q <<- lm(Y ~ ., dat) %>%
+#   broom::tidy() %>%
+#   .[["estimate"]] %>%
+#   .[2]
 r2 <<- lm(Y ~ ., dat) %>% 
   summary() %>% 
   .[["r.squared"]]

### 2020-09-22 R/impute.R

@@ -56,7 +56,7 @@ impute <-
       it = it_nr,
       est = est,
       CIW = ci_up - ci_lo,
-      cov = ci_lo < Q & Q < ci_up,
+      cov = ci_lo < 1 & 1 < ci_up,
       rsq = r_sq
       #qhat = t(qhat),
       #lambda = t(lambda)

### 2020-09-22 R/setup.R

@@ -19,8 +19,7 @@ setup <- function(seed) {
   n_imp <<- 5
   m_mech <<- c("MCAR", "MAR", "MNAR")
   p_inc <<- c(0.05, 0.5, 0.95)
-  thetas <<- c("mu.Y", "mu.X1", "mu.X2", "sigma.Y", "sigma.X1", "sigma.X2", "qhat", "lambda")
-  
+
   # objects to store imputations in
   mids <<- chainmeans <<- chainvars <<- qhats <<- lambdas <<- NULL
   

### 2020-09-22 RunSimulation.R

@@ -56,6 +56,9 @@ parameters <- mus %>%
   dplyr::full_join(qhats) %>% 
   dplyr::full_join(lambdas) 
 
+# get theta names for mapping
+thetas <- names(parameters)[-c(1:5)]
+
 # apply convergence diagnostics to each theta
 convergence_diagnostics <- purrr::map_dfr(1:n_sim, function(ss) {
   purrr::map_dfr(m_mech, function(mm) {
@@ -79,12 +82,6 @@ convergence_diagnostics <- purrr::map_dfr(1:n_sim, function(ss) {
     })
   })
 })
-# thetas %>% 
-#   select(sim, mech, p, m, it, "lambda") %>% 
-#   filter(mech == "MCAR", p == 0.5, sim == 1) %>%  
-#   pivot_wider(names_from = m, values_from = "lambda") %>% 
-#   .[,-c(1:4)] %>% 
-#   convergence()
 
 # summarize results
 conv_results <- convergence_diagnostics %>% 
@@ -96,10 +93,7 @@ results <- map_df(outcomes, ~ {
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)}) %>% 
   full_join(., conv_results)
 
-save(parameters, file = "Data/parameters_cov.Rdata")
-save(outcomes, file = "Data/outcomes_cov.Rdata")
-save(results, file = "Data/results_cov.Rdata")
-
-# save(parameters, file = "Data/parameters.Rdata")
-# save(outcomes, file = "Data/outcomes.Rdata")
-# save(results, file = "Data/results.Rdata")
+save(parameters, file = "Data/raw_parameters.Rdata")
+save(outcomes, file = "Data/raw_outcomes.Rdata")
+save(convergence_diagnostics, file = "Data/diagnostics.Rdata")
+save(results, file = "Data/results.Rdata")

### 2020-09-22 Manuscript.Rmd
New regression coefficient, 100 sims

@@
 ```{r}
-load("~/MissingThePoint/Data/convergence_diagnostics_test.Rdata")
-convergence_diagnostics %>% filter(sim == 1, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line()
-convergence_diagnostics %>% filter(sim == 1, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line()
+# load("~/MissingThePoint/Data/diagnostics.Rdata")
+# load("~/MissingThePoint/Data/raw_outcomes.Rdata")
+sim_nr <- 3
+outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line()
+convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line()
+convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line()
 ```
+
+## Example data
+
+```{r}
+# environment
+library(mice)
+set.seed(123)
+
+# functions
+source("~/MissingeThePoint/R/ComputeDiagnostics.R")
+
+not_all_na <- function(x) {
+  !all(is.na(x))
+}
+
+preprocess_thetas <- function(theta, ext){
+  out <- theta %>% 
+  as.data.frame() %>% 
+  t() %>% 
+  as.data.frame() %>%
+        setNames(paste0(ext, ".", names(.))) %>%
+        cbind(
+          it = 1:n_it,
+          m = rep(1:n_imp, each = n_it),
+          .
+        ) %>% 
+    select_if(not_all_na) %>% 
+    arrange(it)
+  rownames(out) <- NULL
+  return(out)
+}
+
+get_diagnostics <- function(parameters) {
+  thetas <- names(parameters)[-c(1:2)]
+  out <- purrr::map_dfc(thetas, function(tt) {
+    matrix(parameters[[tt]], ncol = n_imp, byrow = T) %>%
+      convergence() %>%
+      select(r.hat.max, ac.max) %>%
+      setNames(., paste0(names(.), ".", tt))
+  }) %>%
+    cbind(it = 1:n_it, .)
+  return(out)
+}
+
+impute_once <- function(iteration, ...) {
+  if(is.null(mids)){
+    mids <<- mice(boys, m = n_imp, maxit = 1, printFlag = FALSE)
+  }
+  else {
+    mids <<- mice.mids(mids, printFlag = FALSE)
+  }
+  mira <- mids %>% mice::lm.mids(age ~ ., data = .) 
+  mipo <- mira %>%
+    mice::pool() %>%
+    summary(., conf.int = TRUE)
+  out <- data.frame(it = iteration, est = mipo[["estimate"]][2])
+  return(out)
+}
+
+# parameters
+n_imp <- 5
+n_it <- 10
+
+# analyses
+mids <- NULL
+ests <- purrr::map_dfr(1:n_it, function(it){impute_once(iteration = it)})
+param <- preprocess_thetas(mids$chainMean, ext = "mu") %>% 
+  dplyr::full_join(
+      preprocess_thetas(mids$chainVar, ext = "sigma")
+  )
+convs <- get_diagnostics(param)
+res <- full_join(ests, convs)
+# plot
+ggplot(conv, aes(x = it, y = ac.max.mu.hgt)) +
+  geom_line()
+
+# more analyses
+imp2 <- mice(boys, m = n_imp, maxit = 1, printFlag = FALSE, seed = 123)
+mice.mids(imp2)
+```

### 2020-09-22 R/impute.R

@@ -26,7 +26,7 @@ impute <-
     
     # analyze and pool imputations
     mira <- mids %>%
-      mice::lm.mids(Y ~ ., .) 
+      mice::lm.mids(Y ~ ., data = .) 
     mipo <- mira %>%
       mice::pool() %>%
       summary(., conf.int = TRUE)

### 2020-09-22 Manuscript.Rmd
Add example of diagnostics with boys data

@@
 # functions
-source("~/MissingeThePoint/R/ComputeDiagnostics.R")
+source("~/MissingThePoint/R/ComputeDiagnostics.R")
 
@@
   if(is.null(mids)){
-    mids <<- mice(boys, m = n_imp, maxit = 1, printFlag = FALSE)
+    mids <<- mice(boys, m = n_imp, maxit = 1, seed = 123, printFlag = FALSE)
   }
@@
     summary(., conf.int = TRUE)
+  mild <- mids %>% mice::complete("all")
+
+  qhat <- 
+      map_dbl(mild, ~ {
+        lm(formula = age ~ ., data = .) %>% .$coefficients %>% .[2]
+      })
+  lambda <- mild %>%
+      purrr::map_dbl(., ~ {
+        dplyr::mutate(., dplyr::across(dplyr::everything(), ~as.numeric(.))) %>% 
+        princomp(., cor = TRUE) %>% .$sdev %>% .[1] %>% . ^ 2 #first eigenvalue of the varcovar matrix
+      })
+  qhats <<- rbind(qhats, data.frame(it = iteration, m = 1:n_imp, qhat = qhat))
+  lambdas <<- rbind(lambdas, data.frame(it = iteration, m = 1:n_imp, lambda = lambda))
+
   out <- data.frame(it = iteration, est = mipo[["estimate"]][2])
@@
 n_imp <- 5
-n_it <- 10
+n_it <- 20
 
 # analyses
-mids <- NULL
+mids <- qhats <- lambdas <- NULL
 ests <- purrr::map_dfr(1:n_it, function(it){impute_once(iteration = it)})
+mids <- mice(boys, m = n_imp, maxit = n_it, seed = 123, printFlag = FALSE)
 param <- preprocess_thetas(mids$chainMean, ext = "mu") %>% 
   dplyr::full_join(
-      preprocess_thetas(mids$chainVar, ext = "sigma")
+      preprocess_thetas(mids$chainVar, ext = "sigma") %>% 
+        dplyr::full_join(qhats) %>% 
+        dplyr::full_join(lambdas)
   )
@@
 res <- full_join(ests, convs)
+
 # plot
-ggplot(conv, aes(x = it, y = ac.max.mu.hgt)) +
+ggplot(res, aes(x = it, y = est)) +
+  geom_line()
+ggplot(res, aes(x = it, y = ac.max.mu.hgt)) +
+  geom_line()
+ggplot(res, aes(x = it, y = ac.max.qhat)) +
+  geom_line()
+ggplot(res, aes(x = it, y = ac.max.lambda)) +
+  geom_line()
+ggplot(res, aes(x = it, y = r.hat.max.mu.hgt)) +
+  geom_line()
+ggplot(res, aes(x = it, y = r.hat.max.qhat)) +
+  geom_line()
+ggplot(res, aes(x = it, y = r.hat.max.lambda)) +
   geom_line()
 
-# more analyses
-imp2 <- mice(boys, m = n_imp, maxit = 1, printFlag = FALSE, seed = 123)
-mice.mids(imp2)
 ```

### 2020-09-23 Manuscript.Rmd
Extend example with plots

--- 2020-09-22 Manuscript.Rmd
@@
 ```{r}
-load("~/MissingThePoint/Data/convergence_diagnostics_test.Rdata")
-convergence_diagnostics %>% filter(sim == 1, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line()
-convergence_diagnostics %>% filter(sim == 1, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line()
+# load("~/MissingThePoint/Data/diagnostics.Rdata")
+# load("~/MissingThePoint/Data/raw_outcomes.Rdata")
+sim_nr <- 3
+outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line()
+convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line()
+convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line()
 ```
+
+## Example data
+
+```{r}
+load("~/MissingThePoint/Data/example.Rdata")
+
+# plot example
+plotfun <- function(v){
+ggplot(example, aes(x = it, y = .data[[v]])) +
+  geom_line() +
+  theme_classic()
+}
+
+plotfun(v="est")
+plotfun(v="ac.max.qhat")
+plotfun(v="ac.max.lambda")
+plotfun(v="r.hat.max.qhat")+list(geom_hline(yintercept = 1.2, color = "grey"))
+plotfun(v="r.hat.max.lambda")+list(geom_hline(yintercept = 1.2, color = "grey"))
+
+load("~/MissingThePoint/Data/example_mids.Rdata")
+plot(mids)
+```

### 2020-09-23 CreateExampleR.R

@@ -1,0 +1,16 @@
+# Plot example of convergence diagnostics and outcomes
+
+# setup
+source("~/MissingThePoint/R/setup_example.R")
+setup_example(seed=11)
+ests <- purrr::map_dfr(1:n_it, function(it){impute_once(iteration = it)})
+param <- preprocess_thetas(mids$chainMean, ext = "mu") %>% 
+  dplyr::full_join(
+    preprocess_thetas(mids$chainVar, ext = "sigma") %>% 
+      dplyr::full_join(qhats) %>% 
+      dplyr::full_join(lambdas)
+  )
+convs <- get_diagnostics(param)
+example <- full_join(ests, convs)
+save(example, file = "Data/example.Rdata")
+save(mids, file = "Data/example_mids.Rdata")

### 2020-09-23 R/setup_example.R

@@ -1,0 +1,80 @@
+# setup environment for example
+
+setup_example <- function(seed){
+  # environment
+  library(mice)
+  library(tidyverse)
+  set.seed(seed)
+  
+  # parameters
+  n_imp <<- 5
+  n_it <<- 25
+  mids <<- qhats <<- lambdas <<- NULL  
+}
+
+# other functions to load
+source("~/MissingThePoint/R/ComputeDiagnostics.R")
+
+not_all_na <- function(x) {
+  !all(is.na(x))
+}
+
+preprocess_thetas <- function(theta, ext){
+  out <- theta %>% 
+    as.data.frame() %>% 
+    t() %>% 
+    as.data.frame() %>%
+    setNames(paste0(ext, ".", names(.))) %>%
+    cbind(
+      it = 1:n_it,
+      m = rep(1:n_imp, each = n_it),
+      .
+    ) %>% 
+    select_if(not_all_na) %>% 
+    arrange(it)
+  rownames(out) <- NULL
+  return(out)
+}
+
+get_diagnostics <- function(parameters) {
+  thetas <- names(parameters)[-c(1:2)]
+  out <- purrr::map_dfc(thetas, function(tt) {
+    matrix(parameters[[tt]], ncol = n_imp, byrow = T) %>%
+      convergence() %>%
+      select(r.hat.max, ac.max) %>%
+      setNames(., paste0(names(.), ".", tt))
+  }) %>%
+    cbind(it = 1:n_it, .)
+  return(out)
+}
+
+impute_once <- function(iteration, ...) {
+  if(is.null(mids)){
+    mids <<- mice(boys, m = n_imp, maxit = 1, printFlag = FALSE)
+  }
+  else {
+    mids <<- mice.mids(mids, printFlag = FALSE)
+  }
+  mira <- mids %>% mice::lm.mids(age ~ ., data = .) 
+  mipo <- mira %>%
+    mice::pool() %>%
+    summary(., conf.int = TRUE)
+  mild <- mids %>% mice::complete("all")
+  
+  qhat <- 
+    map_dbl(mild, ~ {
+      lm(formula = age ~ ., data = .) %>% .$coefficients %>% .[2]
+    })
+  lambda <- mild %>%
+    purrr::map_dbl(., ~ {
+      dplyr::mutate(., dplyr::across(dplyr::everything(), ~as.numeric(.))) %>% 
[…]

### 2020-09-24 Manuscript.Rmd
Sim results with 25/50/75% incomplete instead of 5/50/95

@@
 ```{r setup, include=FALSE}
+# environment
 knitr::opts_chunk$set(echo = TRUE)
 library(tidyverse)
-load("~/MissingThePoint/Data/results_test.Rdata")
+library(mice)
+load("Data/results.Rdata")
 
+# sim results plotting function
+generic_plot <- function(v, name, ...){
+  results %>%
+    mutate(
+      p = factor(p, levels = c(0.05, 0.5, 0.95), labels = c("5%", "50%", "95%")),
+      mech = factor(mech, levels = c("MCAR", "MAR", "MNAR"))
+      ) %>% 
+  ggplot(aes(
+    x=.data[["it"]], 
+    y=.data[[v]], 
+    color = .data[["p"]], 
+    shape = .data[["mech"]], 
+    linetype = .data[["mech"]]
+    )) +
+  #geom_point(na.rm = TRUE) +
+  geom_line(na.rm = TRUE) +
+  labs(
+    x = "Number of iterations",
+    y = name,
+    color = "Incomplete \n cases", 
+    shape = "Missingness \n mechanism", 
+    linetype = "Missingness \n mechanism") +
+  theme_classic() + 
+  scale_color_manual(values = c(
+    #"#009E73", #green
+    "#56B4E9", #blue
+    #"#E69F00", #orange
+    "#F0E442", #yellow
+    "#D55E00"  #red
+  ))
+  # scale_color_manual(values = c(
+  #   # '#228833', #green
+  #   # '#66CCEE', #blue
+  #   '#CCBB44', #yellow
+  #   '#EE6677',  #red~ish
+  #   '#AA3377'  #maroon
+  #   ))
+}
 ```
@@
 
-- In this paper we explore diagnostic methods for MICE, and show which parameter has the best performance --> most informative about convergence
+- In this paper we explore diagnostic methods for MICE, and show which method (parameter + diagnostic + threshold) is the most informative about convergence
 
@@
 
-- What does non-convergence look like? What are the consequences?
+- What does non-convergence look like? What are the consequences? Refer to van Buuren (2018) instead of reproducing it.
 
@@
 ```{r echo=FALSE}
-results %>%
-  ggplot() +
-  #geom_hline(yintercept = 0.3334) +
-  geom_point(aes(x=it, y=est, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=est, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+generic_plot(v="est", name = "Regression estimate") + 
+  list(geom_hline(yintercept = 1, color = "grey"))
 
@@
 
-```{r}
-results %>% #filter(it>6) %>% 
-  ggplot() +
-  geom_hline(yintercept = 0.95, size = 2, color = "grey") +
-  geom_point(aes(x=it, y=cov, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=cov, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="cov", name = "Coverage rate") + 
+  list(geom_hline(yintercept = 0.95, color = "grey"))
 
@@
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=CIW, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=CIW, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="CIW", name = "CI width")
 
@@
 
-```{r}
-results %>% 
-  # filter(it>6) %>% 
-  ggplot() +
-  geom_point(aes(x=it, y=rsq, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=rsq, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="rsq", name = "Coefficient of determination (r^2)") +
+  list(geom_hline(yintercept = 0.5, color = "grey"))
 
@@
 
-Magnitude of the bias in R^2 perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for p=0.95, MNAR, but stable from it=5.
+Magnitude of the bias in r^2 perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for p=0.95, MNAR, but stable from it=5.
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=ac.max.mu.Y, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=ac.max.mu.Y, color = mech, linetype = as.factor(p))) +
-  theme_classic()
-```
-
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=r.hat.max.mu.Y, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=r.hat.max.mu.Y, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="ac.max.mu.Y", name = "AC chain means")
 
@@
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=ac.max.sigma.Y, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=ac.max.sigma.Y, color = mech, linetype = as.factor(p))) +
-  theme_classic()
-```
+This plot above and the next three plots show autocorrelations with different thetas
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=r.hat.max.sigma.Y, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=r.hat.max.sigma.Y, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="ac.max.sigma.Y", name = "AC chain variances")
 
@@
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=ac.max.qhat, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=ac.max.qhat, color = mech, linetype = as.factor(p))) +
-  theme_classic()
-```
-
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=r.hat.max.qhat, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=r.hat.max.qhat, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="ac.max.qhat", name = "AC scientific estimate")
 
@@
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=ac.max.lambda, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=ac.max.lambda, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="ac.max.lambda", name = "AC novel theta")
+
 ```
 
-```{r}
-results %>%
-  ggplot() +
-  geom_point(aes(x=it, y=r.hat.max.lambda, color = mech, shape = as.factor(p))) +
-  geom_line(aes(x=it, y=r.hat.max.lambda, color = mech, linetype = as.factor(p))) +
-  theme_classic()
+```{r echo=FALSE}
+generic_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
+
+```
+
+This plot above and the next three plots show r hat values with different thetas
+
+```{r echo=FALSE}
+generic_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
+
+```
+
+```{r echo=FALSE}
+generic_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
+
+```
+
+```{r echo=FALSE}
+generic_plot(v="r.hat.max.lambda", name = "R hat novel theta")
 
@@
 
-## Check if heuristic for AC works
-```{r}
-# load("~/MissingThePoint/Data/diagnostics.Rdata")
-# load("~/MissingThePoint/Data/raw_outcomes.Rdata")
-sim_nr <- 3
-outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line()
-convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line()
-convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line()
-```
+<!-- ## Check if heuristic for AC works -->
+<!-- ```{r} -->
+<!-- load("Data/diagnostics.Rdata") -->
+<!-- load("Data/raw_outcomes.Rdata") -->
+<!-- sim_nr <- 3 -->
+<!-- outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line() -->
+<!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line() -->
+<!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line() -->
+<!-- ``` -->
 
@@
 ```{r}
-load("~/MissingThePoint/Data/example.Rdata")
+# load data
+load("Data/example.Rdata")
+load("Data/example_mids.Rdata")
 
-# plot example
-plotfun <- function(v){
-ggplot(example, aes(x = it, y = .data[[v]])) +
+# plot each var of example data
+plotfun <- function(v, ...){
+ggplot(example, aes(x = .data[["it"]], y = .data[[v]])) +
   geom_line() +
@@
 
-load("~/MissingThePoint/Data/example_mids.Rdata")
 plot(mids)

### 2020-09-24 R/setup.R

@@ -18,7 +18,7 @@ setup <- function(seed) {
   n_cases <<- 1000
   n_imp <<- 5
   m_mech <<- c("MCAR", "MAR", "MNAR")
-  p_inc <<- c(0.05, 0.5, 0.95)
+  p_inc <<- c(0.25, 0.5, 0.75)
 
   # objects to store imputations in
   mids <<- chainmeans <<- chainvars <<- qhats <<- lambdas <<- NULL

### 2020-09-24 RunSimulation.R

@@ -93,7 +93,7 @@ results <- map_df(outcomes, ~ {
 }) %>% aggregate(. ~ it + p + mech, data = ., function(x){mean(x, na.rm = TRUE)}) %>% 
   full_join(., conv_results)
 
-save(parameters, file = "Data/raw_parameters.Rdata")
-save(outcomes, file = "Data/raw_outcomes.Rdata")
-save(convergence_diagnostics, file = "Data/diagnostics.Rdata")
-save(results, file = "Data/results.Rdata")
+save(parameters, file = "Data/raw_parameters_25_75.Rdata")
+save(outcomes, file = "Data/raw_outcomes_25_75.Rdata")
+save(convergence_diagnostics, file = "Data/diagnostics_25_75.Rdata")
+save(results, file = "Data/results_25_75.Rdata")

### 2020-09-24 Manuscript.Rmd
Interpret rhat curves

@@
     )) +
-  #geom_point(na.rm = TRUE) +
   geom_line(na.rm = TRUE) +
@@
 ```{r echo=FALSE}
-generic_plot(v="est", name = "Regression estimate") + 
+generic_plot(v = "est", name = "Regression estimate") + 
   list(geom_hline(yintercept = 1, color = "grey"))
@@
 ```{r echo=FALSE}
-generic_plot(v="cov", name = "Coverage rate") + 
+generic_plot( v ="cov", name = "Coverage rate") + 
   list(geom_hline(yintercept = 0.95, color = "grey"))
+
+# generic_plot( v ="cov", name = "Coverage rate") + 
+#   list(geom_hline(yintercept = 0.95, color = "grey")) +
+#   ylim(0.85, 1)
 
@@
 
+These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??
+
 ## Discussion

### 2020-09-29 Manuscript.Rmd
Edit intro

@@
 
-Missing data poses a ubiquitous threat against unbiased, confidence-valid statistical inferences.
-A popular tool to accommodate missing data is iterative imputation.
-With techniques like multiple imputation (Rubin, 1987), missing data points are imputed (i.e., 'filled in')
+Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences.
+A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset.
+Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inferences under many missing data circumstances (Rubin, 1987; van Buuren, 2018).
+Most imputation software packages rely on iterative algorithms to obtain imputed values.
+
+With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration.
+This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+Are the imputations then to be trusted?
+And can we rely on the inference obtained using the imputed data?
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2).
+While there is no scientific consensus on how to evaluate the convergence of imputation algorithms (Zhu & Raghunathan, 2015; Takahashi, 2017), the current practice is to visually inspect imputations for signs of non-convergence.
+
+Identifying non-convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence (Van Buuren, 2018, 6.5.2).
+Therefore, a quantitative diagnostic method to identify non-convergence would be preferred.
+
+In this paper we explore diagnostic methods for iterative imputation algorithms.
+For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package (REF) in `R` (REF).
+We consider two tested non-convergence identifiers for MCMC algorithms: autocorrelation (conform Lynch, 2007, p.
+147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p.
+5).
+MICE, and show which method (parameter + diagnostic + threshold) is the most informative about convergence
 
@@
 
+methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --> will always indicate conv slower than the est
+
+calc rhat and cov for each block of 5 it
+
+disc: conf val is likely to happen much sooner
+
+as soon as we start downweighting the first few it from the calc, the memory effect would
+
 ## Example data
@@
 plotfun(v="est")
-plotfun(v="ac.max.qhat")
+plotfun(v="ac.max.qhat") #+ list(scale_y_continuous(limits = c(0.65,1)))
 plotfun(v="ac.max.lambda")
-plotfun(v="r.hat.max.qhat")+list(geom_hline(yintercept = 1.2, color = "grey"))
-plotfun(v="r.hat.max.lambda")+list(geom_hline(yintercept = 1.2, color = "grey"))
+plotfun(v="r.hat.max.qhat")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+plotfun(v="r.hat.max.lambda")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
 

### 2020-10-12 Manuscript.Rmd
Add automatically updated references

@@
 output: html_document
+bibliography: ref.bib
 ---
@@
 A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset.
-Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inferences under many missing data circumstances (Rubin, 1987; van Buuren, 2018).
-Most imputation software packages rely on iterative algorithms to obtain imputed values.
+Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances (Rubin, 1987; van Buuren, 2018).
+To obtain imputations, most imputation software packages rely on iterative algorithms.
 
@@
 In this paper we explore diagnostic methods for iterative imputation algorithms.
-For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package (REF) in `R` (REF).
-We consider two tested non-convergence identifiers for MCMC algorithms: autocorrelation (conform Lynch, 2007, p.
+For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+We consider two non-convergence identifiers for iterative algorithms: autocorrelation (conform Lynch, 2007, p.
 147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p.
 5).
-MICE, and show which method (parameter + diagnostic + threshold) is the most informative about convergence
+Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm.
+
+We propose a novel multivariate parameter to check for non-convergence in iterative algorithms.
+Our aim is to show which method (parameter + diagnostic + threshold) is the most informative about non-convergence in the iterative imputation algorithm implemented in `mice`.
 
@@
 
-as soon as we start downweighting the first few it from the calc, the memory effect would
+as soon as we start down-weighting the first few it from the calc, the memory effect would
 
@@
 plotfun(v="ac.max.lambda")
+plotfun(v="r.hat.max.sigma.hgt")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
 plotfun(v="r.hat.max.qhat")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
@@
 ```
+
+# References

### 2020-10-12 R/ComputeDiagnostics.R

@@ -9,8 +9,8 @@ source('R/ComputeAC.R')
 convergence <- function(x, include_acf = FALSE) {
   # input: object with imputation chain(s), e.g. chain mean, choice of calculating default AC
   # output: AC values at each iteration, computed manually, if necessary also using stats:acf() 
-  out <-  left_join(rhat_adapted(x), ac_adapted(x), by = "iteration") 
-  if(include_acf){ out <- left_join(out, ac_adapted(x, "acf"), by = "iteration")
+  out <-  full_join(rhat_adapted(x), ac_adapted(x), by = "iteration") 
+  if(include_acf){out <- full_join(out, ac_adapted(x, "acf"), by = "iteration")
   }
   out <- out %>% dplyr::rename(it = iteration)
   return(out)

### 2020-10-15 Manuscript.Rmd
methods section manuscript (before updating r and rstudio)

@@
 library(mice)
-load("Data/results.Rdata")
+load("Data/results_25_75.Rdata")
 
@@
     mutate(
-      p = factor(p, levels = c(0.05, 0.5, 0.95), labels = c("5%", "50%", "95%")),
+      p = factor(p, levels = c(0.25, 0.5, 0.75), labels = c("25%", "50%", "75%")),
       mech = factor(mech, levels = c("MCAR", "MAR", "MNAR"))
@@
 We propose a novel multivariate parameter to check for non-convergence in iterative algorithms.
-Our aim is to show which method (parameter + diagnostic + threshold) is the most informative about non-convergence in the iterative imputation algorithm implemented in `mice`.
+Our aim is to show which method is the most informative about non-convergence in iterative imputation.
+[explain that method = parameter + diagnostic + interpretation]
 
-- Missing data has to be accommodated, usually filled in
+<!-- - Missing data has to be accommodated, usually filled in -->
 
-- Filling in is algorithmic, which relies on convergence
+<!-- - Filling in is algorithmic, which relies on convergence -->
 
-- Diagnosing convergence is difficult, both visually and diagnostically
+<!-- - Diagnosing convergence is difficult, both visually and diagnostically -->
 
-- In this paper we explore diagnostic methods for MICE, and show which method (parameter + diagnostic + threshold) is the most informative about convergence
+<!-- - In this paper we explore diagnostic methods for MICE, and show which method (parameter + diagnostic + threshold) is the most informative about convergence  -->
 
@@
 
-- What does non-convergence look like? What are the consequences? Refer to van Buuren (2018) instead of reproducing it.
+- Iterative imputation is a sort of MCMC algorithm, so it makes sense to investigate non-convergence in a similar fashion to typical MCMC algorithms.
 
--
+- What does non-convergence look like? --> non-stat + non-mix in a certain parameter, e.g., chain means.
+
+- What are the consequences? --> bias, under-coverage --> Refer to van Buuren (2018) instead of reproducing this
+
+- Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g. N_eff is estimated using varcov matrix of the set of samples, while this is not the case in it im per se.
+
+- We consider autocorr and rhat, and monitor four parameters: chain means, chain variances, the scientific estimand, and lambda.
 
@@
 
-- 3x3 missingness conditions: MCAR, MAR, MNAR + 5, 50, 95% of cases incomplete
+We investigate non-convergence of the MICE algorithm through model-based simulation in R (version 4.0.2; R Core Team 2020).
+The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below
 
-- outcomes: estimated regression coefficient, CIW, coverage, R2 --> multivariate is key, because univariate convergence does not guarantee multivariate convergence
+```
+# simulation set-up (pseudo-code)
+for each simulation repetition (1:1000)
+  1. simulate complete data
+  for each missingness condition (1:9)
+    2. create missingness
+    for each iteration (1:50)
+     3. impute missingness
+     4. perform analysis of scientific interest
+     5. compute non-convergence diagnostics 
+     6. pool results across imputations
+     7. compute performance measures
+ 8. combine outcomes of all missingness conditions
+9. aggregate outcomes of all simulation repetitions
+```
 
-- convergence parameters: chain means, chain variances, scientific parameter, lambda
+### Aims
+Our aim is to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice`.
+To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths.
 
-- diagnostics: rhat, ac
+### Data generating mechanism
+Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases. Let the complete set be three multivariate normally distributed random variables
+
+$$
+\begin{pmatrix}
+Y\\
+X_1\\
+X_2
+\end{pmatrix}
+\sim \mathcal{N}
+\begin{bmatrix}
+\begin{pmatrix}
+0\\
+0\\
+0
+\end{pmatrix},
+\begin{pmatrix}
+2     &       &  \\
+0.5   & 1     &  \\
+-0.5  & 0.5   & 1
+\end{pmatrix}
+\end{bmatrix}.
+$$
+The complete set is amputed according to a $3\times3$ factorial design, consisting of three missingness mechanisms and three proportions of incomplete cases.
+[Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences)]
+
+### Estimands
+We perform multiple linear regression to predict outcome variable $Y$: $Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2$, where $Y'$ is the expected value of the outcome.
+Our estimands are regression coefficient $\beta_1$, and the coefficient of determination, $r^2$.
+
+### Methods
+We use a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.
+
+### Performance measures
+As recommended by @buur18, our performance measures are bias, empirical coverage rate and confidence interval width of the estimands.
+
+<!-- - outcomes: estimated regression coefficient, CIW, coverage, r2 multivariate is key, because univariate convergence does not guarantee multivariate convergence -->
+
+<!-- - convergence parameters: chain means, chain variances, scientific parameter, lambda -->
+
+<!-- - diagnostics: rhat, ac -->
 
@@
 
-Magnitude of the bias in r^2 perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for p=0.95, MNAR, but stable from it=5.
+Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=5.
 
@@
 
-# References
+# References (incomplete)

### 2020-10-15 Manuscript.Rmd
Restructure results section

@@
       p = factor(p, levels = c(0.25, 0.5, 0.75), labels = c("25%", "50%", "75%")),
-      mech = factor(mech, levels = c("MCAR", "MAR", "MNAR"))
+      mech = factor(mech, levels = c("MCAR", "MAR", "MNAR")),
+      est = est - 1,
+      rsq = rsq - 0.5
       ) %>% 
@@
 
-- What does non-convergence look like? --> non-stat + non-mix in a certain parameter, e.g., chain means.
+-   What does non-convergence look like? --\> non-stat + non-mix in a certain parameter, e.g., chain means.
 
-- What are the consequences? --> bias, under-coverage --> Refer to van Buuren (2018) instead of reproducing this
+-   What are the consequences? --\> bias, under-coverage --\> Refer to van Buuren (2018) instead of reproducing this
 
-- Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g. N_eff is estimated using varcov matrix of the set of samples, while this is not the case in it im per se.
+-   Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g. N\_eff is estimated using varcov matrix of the set of samples, while this is not the case in it im per se.
 
@@
 
-We investigate non-convergence of the MICE algorithm through model-based simulation in R (version 4.0.2; R Core Team 2020).
-The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below
+<!-- The aim of the simulation study is to determine the impact of non-convergence on the validity of statistical inferences, and to assess whether non-convergence may be detected using several diagnostic methods.
+Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions (nsim = 1000).
+-->
 
-```
+We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.0.2; R Core Team 2020).
+We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+
     # simulation set-up (pseudo-code)
@@
     9. aggregate outcomes of all simulation repetitions
-```
 
 ### Aims
-Our aim is to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice`.
-To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths.
+
+Our aim is to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice].
+Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
+To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 50$).
+We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].
 
 ### Data generating mechanism
-Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases. Let the complete set be three multivariate normally distributed random variables
+
+Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases. Let the complete set be three multivariately normal random variables
 
@@
 $$
+
 The complete set is amputed according to a $3\times3$ factorial design, consisting of three missingness mechanisms and three proportions of incomplete cases.
@@
 ### Estimands
-We perform multiple linear regression to predict outcome variable $Y$: $Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2$, where $Y'$ is the expected value of the outcome.
-Our estimands are regression coefficient $\beta_1$, and the coefficient of determination, $r^2$.
+
+We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice].
+On each imputed dataset, we perform multiple linear regression to predict outcome variable $Y$ from the other two variables
+
+$$
+Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
+$$
+where $Y'$ is the expected value of the outcome. Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $r^2$ that we obtain after pooling across the imputations.
 
 ### Methods
+
 We use a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.
@@
 ### Performance measures
-As recommended by @buur18, our performance measures are bias, empirical coverage rate and confidence interval width of the estimands.
 
-<!-- - outcomes: estimated regression coefficient, CIW, coverage, r2 multivariate is key, because univariate convergence does not guarantee multivariate convergence -->
+As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($\bar{Q}_{LL} - \bar{Q}_{UL}$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands.
 
-<!-- - convergence parameters: chain means, chain variances, scientific parameter, lambda -->
+## Results
 
-<!-- - diagnostics: rhat, ac -->
+We present the results of our simulation study visually because of the many simulation conditions.
+Within each plot, the results are split according to the missingness conditions (missingness mechanisms as line types, and proportion of incomplete cases as colors).
 
-## Some results
+### Non-Convergence Identifiers
+
+The next four plots show the autocorrelations in each parameter plotted against the number of iterations.
 
 ```{r echo=FALSE}
-generic_plot(v = "est", name = "Regression estimate") + 
-  list(geom_hline(yintercept = 1, color = "grey"))
-
+p <- generic_plot(v="ac.max.mu.Y", name = "AC chain means")
+p %>% plotly::ggplotly()
 ```
 
-Bias in estimate depends mostly on the amount of missingness (less info to estimate the relation??), but overall MNAR performs the worst, which is stable after it=5.
-
-Update with different CI computation, see https://stefvanbuuren.name/fimd/sec-evaluation.html#sec:quantifyingbias
+This plot shows that irrespective of the missingness condition, autocorrelation in the chain means decreases rapidly until $it \geq 6$.
+This means that there is some initial trending within chains, but the stationarity does not improve substantively after the first few iterations.
 
 ```{r echo=FALSE}
-generic_plot( v ="cov", name = "Coverage rate") + 
+p <- generic_plot(v="ac.max.sigma.Y", name = "AC chain variances")
+p %>% plotly::ggplotly() 
+```
+
+Autocorrelations in the chain variances do not show a systematic effect of the missingness conditions either.
+The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous.
+
+```{r echo=FALSE}
+p <- generic_plot(v="ac.max.qhat", name = "AC scientific estimate")
+p %>% plotly::ggplotly() 
+```
+
+```{r echo=FALSE}
+p <- generic_plot(v="ac.max.lambda", name = "AC novel theta")
+p %>% plotly::ggplotly() 
+```
+
+The next four plots show r hat values with different thetas
+
+```{r echo=FALSE}
+p <- generic_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
+p %>% plotly::ggplotly() 
+```
+
+```{r echo=FALSE}
+p <- generic_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
+p %>% plotly::ggplotly() 
+```
+
+```{r echo=FALSE}
+p <- generic_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
+p %>% plotly::ggplotly() 
+```
+
+```{r echo=FALSE}
+p <- generic_plot(v="r.hat.max.lambda", name = "R hat novel theta")
+p %>% plotly::ggplotly() 
+```
+
+These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??
+
+### Performance Measures
+
+```{r echo=FALSE}
+p <- generic_plot(v = "est", name = "Bias in regression estimate") + 
+  list(geom_hline(yintercept = 0, color = "grey")) 
+p %>% plotly::ggplotly()
+```
+
+Bias in estimate depends mostly on the amount of missingness (less info to estimate the relation??). Overall 75% incomplete cases performs the worst, but is stable for $t \geq 6$.
+
+<!-- Update with different CI computation, see <https://stefvanbuuren.name/fimd/sec-evaluation.html#sec:quantifyingbias> -->
+
+```{r echo=FALSE}
+p <- generic_plot( v ="cov", name = "Coverage rate") + 
   list(geom_hline(yintercept = 0.95, color = "grey"))
+p %>% plotly::ggplotly()
 
-# generic_plot( v ="cov", name = "Coverage rate") + 
+# p <- generic_plot( v ="cov", name = "Coverage rate") + 
 #   list(geom_hline(yintercept = 0.95, color = "grey")) +
@@
 
-Cov is more or less fine, except for MNAR with only 0.05% incomplete cases --> due to bias in the estimate in combination with very narrow CI (see CIW!). However, all stable after it=3??
+Cov is more or less fine, except for MNAR with only 0.05% incomplete cases --\> due to bias in the estimate in combination with very narrow CI (see CIW!). However, all stable after it=3??
 
 ```{r echo=FALSE}
-generic_plot(v="CIW", name = "CI width")
-
+p <- generic_plot(v="CIW", name = "CI width") 
+p %>% plotly::ggplotly()
 ```
@@
 ```{r echo=FALSE}
-generic_plot(v="rsq", name = "Coefficient of determination (r^2)") +
-  list(geom_hline(yintercept = 0.5, color = "grey"))
-
+p <- generic_plot(v="rsq", name = "Bias in coefficient of determination") +
+  list(geom_hline(yintercept = 0, color = "grey"))
+p %>% plotly::ggplotly()
 ```
@@
 
-```{r echo=FALSE}
-generic_plot(v="ac.max.mu.Y", name = "AC chain means")
+## Discussion
 
-```
-
-This plot above and the next three plots show autocorrelations with different thetas
-
-```{r echo=FALSE}
-generic_plot(v="ac.max.sigma.Y", name = "AC chain variances")
-
-```
-
-```{r echo=FALSE}
-generic_plot(v="ac.max.qhat", name = "AC scientific estimate")
-
-```
-
-```{r echo=FALSE}
-generic_plot(v="ac.max.lambda", name = "AC novel theta")
-
-```
-
-```{r echo=FALSE}
-generic_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
-
-```
-
-This plot above and the next three plots show r hat values with different thetas
-
-```{r echo=FALSE}
-generic_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
-
-```
-
-```{r echo=FALSE}
-generic_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
-
-```
-
-```{r echo=FALSE}
-generic_plot(v="r.hat.max.lambda", name = "R hat novel theta")
-
-```
-
-These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??
-
-## Discussion
+- need for multivar parameters!!
 
@@
 <!-- ## Check if heuristic for AC works -->
+
 <!-- ```{r} -->
+
 <!-- load("Data/diagnostics.Rdata") -->
+
 <!-- load("Data/raw_outcomes.Rdata") -->
+
 <!-- sim_nr <- 3 -->
+
 <!-- outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line() -->
+
 <!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line() -->
+
 <!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line() -->
+
 <!-- ``` -->
 
-methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --> will always indicate conv slower than the est
+methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate conv slower than the est
 

### 2020-10-16 Manuscript.Rmd
Interpret ac curves

@@
 
-    # simulation set-up (pseudo-code)
+Algorithm 1: simulation set-up (pseudo-code)
+
     for each simulation repetition (1:1000)
@@
           4. perform analysis of scientific interest
-         5. compute non-convergence diagnostics 
+          5. apply non-convergence identifiers
           6. pool results across imputations
           7. compute performance measures
-     8. combine outcomes of all missingness conditions
-    9. aggregate outcomes of all simulation repetitions
+        8. combine outcomes from all iterations
+      9. combine outcomes from all missingness conditions
+    10. aggregate outcomes from all simulation repetitions
 
@@
 Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
-To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 50$).
+To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$).
 We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].
@@
 
-Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases. Let the complete set be three multivariately normal random variables
+Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases, before inducing missingness. Let the complete set be three multivariately normal random variables
 
@@
 
-The complete set is amputed according to a $3\times3$ factorial design, consisting of three missingness mechanisms and three proportions of incomplete cases.
-[Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences)]
+The complete set is amputed according to the nine missingness conditions.
+We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.
+[Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences).
+And add the `ampute` function?]
 
@@
 $$
-where $Y'$ is the expected value of the outcome. Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $r^2$ that we obtain after pooling across the imputations.
+where $Y'$ is the expected value of the outcome.
+Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $r^2$ that we obtain after pooling the regression results across the imputations.
 
@@
 Autocorrelations in the chain variances do not show a systematic effect of the missingness conditions either.
-The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous.
+The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous, but roughly when $it \geq 10$.
 
@@
 
+We see that the autocorrelations in the scientific estimates are more pronounced than in the univariate parameters (chain means and chain variances), especially when the proportion of missing cases is moderate to extreme.
+We observe the worst stationarity in conditions with 75% of cases having missing values.
+There is no clear effect of the missingness mechanisms.
+Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$.
+
 ```{r echo=FALSE}
@@
 
-The next four plots show r hat values with different thetas
+Autocorrelations in the novel parameter show a similar trend to the autocorrelations in the scientific estimates. Around $it \geq 20$ is required before trending in this parameter diminishes.
+
+The next four plots show the $\widehat{R}$ in the different parameters.
 

### 2020-10-16 Manuscript.Rmd
Interpret rhats

@@
 
+In this plot we observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations---with the exception of an initial increase when $3 < it < 5$.
+There is no apparent differentiation between the missingness conditions.
+The mixing between chain means generally improves until $it \geq 30$ to $40$.
+
 ```{r echo=FALSE}
@@
 ```
+
+The mixing between chain variances mimics the mixing between chain means almost perfectly. Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $it \geq 30$.
 
@@
 
+With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again.
+We do, however, see some differentiation between missingness conditions.
+Conditions with 75% of cases being incomplete show more extreme non-mixing.
+The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.
+
 ```{r echo=FALSE}
@@
 
-These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??
+$\widehat{R}$-values in the chains of the novel parameter show a trend similar to the chain means and chain variances. [add something about conditions??]
+
+[These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??]
 

### 2020-10-16 Manuscript.Rmd
Convert to pdf

@@
 date: "7-9-2020"
-output: html_document
+output:
+  html_document: default
+  pdf_document: default
 bibliography: ref.bib
@@
 
-Algorithm 1: simulation set-up (pseudo-code)
+*Algorithm 1: simulation set-up (pseudo-code)*
 
@@
 p <- generic_plot(v="ac.max.mu.Y", name = "AC chain means")
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 ```
@@
 p <- generic_plot(v="ac.max.sigma.Y", name = "AC chain variances")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
 p <- generic_plot(v="ac.max.qhat", name = "AC scientific estimate")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
 p <- generic_plot(v="ac.max.lambda", name = "AC novel theta")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
 p <- generic_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
 p <- generic_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
 p <- generic_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
 p <- generic_plot(v="r.hat.max.lambda", name = "R hat novel theta")
-p %>% plotly::ggplotly() 
+p # %>% plotly::ggplotly() 
 ```
@@
   list(geom_hline(yintercept = 0, color = "grey")) 
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 ```
@@
   list(geom_hline(yintercept = 0.95, color = "grey"))
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 
@@
 p <- generic_plot(v="CIW", name = "CI width") 
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 ```
@@
   list(geom_hline(yintercept = 0, color = "grey"))
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 ```

### 2020-10-19 Manuscript.Rmd
Clarify results and grammar check

--- 2020-10-16 Manuscript.Rmd
@@
 date: "7-9-2020"
-output: html_document
+output:
+  word_document: default
+  pdf_document: default
+  html_document: default
 bibliography: ref.bib
@@
 # sim results plotting function
-generic_plot <- function(v, name, ...){
+simulation_plot <- function(v, name, ...){
   results %>%
@@
 
+*Notation in this manuscript: *
+
+- square brackets are comments for myself/stuff I need to review
+
+- bullet points are things I need to expand
+
 ## Abstract
@@
 While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
-Our study provides insight into identifying non-convergence in iterative imputation algorithms, since there is no consensus on how to evaluate the convergence properties currently.
+Our study provides insight into identifying non-convergence in iterative imputation algorithms since there is no consensus on how to evaluate the convergence properties currently.
 We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
@@
 
-In this paper we explore diagnostic methods for iterative imputation algorithms.
+In this paper, we explore diagnostic methods for iterative imputation algorithms.
 For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
@@
 We propose a novel multivariate parameter to check for non-convergence in iterative algorithms.
-Our aim is to show which method is the most informative about non-convergence in iterative imputation.
-[explain that method = parameter + diagnostic + interpretation]
+We aim to show which method is the most informative about non-convergence in iterative imputation [explain that method = parameter + diagnostic + interpretation].
 
@@
 
--   What does non-convergence look like? --\> non-stat + non-mix in a certain parameter, e.g., chain means.
+-   What does non-convergence look like? --\> non-stationarity + non-mixing in a certain parameter, e.g., chain means.
 
--   What are the consequences? --\> bias, under-coverage --\> Refer to van Buuren (2018) instead of reproducing this
+-   What are the consequences? --\> bias, under-coverage --\> refer to van Buuren (2018) instead of reproducing this
 
--   Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g. N\_eff is estimated using varcov matrix of the set of samples, while this is not the case in it im per se.
+<!-- -   Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g.
+N\_eff is estimated using the variance-covariance matrix of the set of samples, while this is not the case in iterative imputation per se.
+-->
 
--   We consider autocorr and rhat, and monitor four parameters: chain means, chain variances, the scientific estimand, and lambda.
+-   We consider autocorrelation and rhat, and monitor four parameters: chain means, chain variances, a scientific estimate, and lambda.
 
-## Simulation
+- Explain lambda
+
+## Simulation Set-Up
 
@@
 
-Algorithm 1: simulation set-up (pseudo-code)
+**Algorithm 1: simulation set-up (pseudo-code)**
 
@@
 
-Our aim is to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice].
+We aim to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice].
 Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
@@
 
-Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases, before inducing missingness. Let the complete set be three multivariately normal random variables
+Data are generated in each simulation repetition for a complete set of $n_{obs} = 1000$ cases (i.e., before inducing missingness). We define three multivariately normal random variables, let
 
@@
 
-The complete set is amputed according to the nine missingness conditions.
+[Add mvtnorm package here??] The complete set is amputed according to nine missingness conditions.
 We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.
@@
 
-We use a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.
+We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.
 
@@
 
-As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($\bar{Q}_{LL} - \bar{Q}_{UL}$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands.
+As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($E(\bar{Q}_{LL} - \bar{Q}_{UL})$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands [or just of the regression estimate??
+otherwise add the ciw and cov of r2 too!!
+and check if definition of ciw is correct].
 
-## Results
+## Simulation Results
 
-We present the results of our simulation study visually because of the many simulation conditions.
-Within each plot, the results are split according to the missingness conditions (missingness mechanisms as line types, and proportion of incomplete cases as colors).
+The following plots provide an overview of the results for each method and each performance measure, contrasted to the number of iteration in the imputation algorithm.
+Within the plots, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
 
-### Non-Convergence Identifiers
+### Diagnostic Methods
 
-The next four plots show the autocorrelations in each parameter plotted against the number of iterations.
+<!-- The next four plots show the autocorrelations in each parameter plotted against the number of iterations.  -->
 
 ```{r echo=FALSE}
-p <- generic_plot(v="ac.max.mu.Y", name = "AC chain means")
-p %>% plotly::ggplotly()
+p <- simulation_plot(v="ac.max.mu.Y", name = "Autocorrelation chain means")
+p # %>% plotly::ggplotly()
 ```
 
-This plot shows that irrespective of the missingness condition, autocorrelation in the chain means decreases rapidly until $it \geq 6$.
-This means that there is some initial trending within chains, but the stationarity does not improve substantively after the first few iterations.
+In this plot, our diagnostic method is the amount of autocorrelation in the chain means.
+We can interpret this as the extent to which the average imputed value trends across iterations.
+We see that autocorrelation in the chain means rapidly decreases until $it \geq 6$.
+This means that there is some initial trending within chains, but stationarity does not improve substantively after the first few iterations.
+These results hold irrespective of the missingness condition.
 
 ```{r echo=FALSE}
-p <- generic_plot(v="ac.max.sigma.Y", name = "AC chain variances")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="ac.max.sigma.Y", name = "Autocorrelation chain variances")
+p # %>% plotly::ggplotly() 
 ```
 
-Autocorrelations in the chain variances do not show a systematic effect of the missingness conditions either.
-The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous, but roughly when $it \geq 10$.
+Autocorrelation in the chain variances shows something similar.
+The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally when $it \geq 10$.
+We do not observe a systematic effect of the missingness conditions here either.
 
 ```{r echo=FALSE}
-p <- generic_plot(v="ac.max.qhat", name = "AC scientific estimate")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="ac.max.qhat", name = "Autocorrelation scientific estimate")
+p # %>% plotly::ggplotly() 
 ```
 
-We see that the autocorrelations in the scientific estimates are more pronounced than in the univariate parameters (chain means and chain variances), especially when the proportion of missing cases is moderate to extreme.
-We observe the worst stationarity in conditions with 75% of cases having missing values.
+There is more autocorrelation in the scientific estimates than in the univariate parameters (chain means and chain variances).
+We observe the highest autocorrelations in conditions where 75% of cases are incomplete.
+Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$.
 There is no clear effect of the missingness mechanisms.
-Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$.
 
 ```{r echo=FALSE}
-p <- generic_plot(v="ac.max.lambda", name = "AC novel theta")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="ac.max.lambda", name = "Autocorrelation novel theta")
+p # %>% plotly::ggplotly() 
 ```
 
-Autocorrelations in the novel parameter show a similar trend to the autocorrelations in the scientific estimates. Around $it \geq 20$ is required before trending in this parameter diminishes.
+The autocorrelation in the novel parameter exhibits a similar trend to the autocorrelation in the scientific estimates. Trending in this parameter diminishes when $it \geq 20$.
 
-The next four plots show the $\widehat{R}$ in the different parameters.
+<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->
 
 ```{r echo=FALSE}
-p <- generic_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
+p # %>% plotly::ggplotly() 
 ```
 
-In this plot we observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations---with the exception of an initial increase when $3 < it < 5$.
+We observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations.
+An exception to this observation is the initial increase when $3 < it < 5$.
+According to these results, the mixing between chain means generally improves until $it \geq 30$ to $40$.
 There is no apparent differentiation between the missingness conditions.
-The mixing between chain means generally improves until $it \geq 30$ to $40$.
 
 ```{r echo=FALSE}
-p <- generic_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
+p # %>% plotly::ggplotly() 
 ```
@@
 ```{r echo=FALSE}
-p <- generic_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
+p # %>% plotly::ggplotly() 
 ```
@@
 We do, however, see some differentiation between missingness conditions.
-Conditions with 75% of cases being incomplete show more extreme non-mixing.
+Conditions where 75% of the cases are incomplete show more extreme non-mixing.
 The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.
@@
 ```{r echo=FALSE}
-p <- generic_plot(v="r.hat.max.lambda", name = "R hat novel theta")
-p %>% plotly::ggplotly() 
+p <- simulation_plot(v="r.hat.max.lambda", name = "R hat novel theta")
+p # %>% plotly::ggplotly() 
 ```
@@
 
-[These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??]
+[These rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]
 
@@
 ```{r echo=FALSE}
-p <- generic_plot(v = "est", name = "Bias in regression estimate") + 
+p <- simulation_plot(v = "est", name = "Bias in regression estimate") + 
   list(geom_hline(yintercept = 0, color = "grey")) 
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 ```
 
-Bias in estimate depends mostly on the amount of missingness (less info to estimate the relation??). Overall 75% incomplete cases performs the worst, but is stable for $t \geq 6$.
+We see that within a few iterations the average bias approaches zero.
+When $it \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
 
@@
 ```{r echo=FALSE}
-p <- generic_plot( v ="cov", name = "Coverage rate") + 
+p <- simulation_plot( v ="cov", name = "Coverage rate") + 
   list(geom_hline(yintercept = 0.95, color = "grey"))
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 
-# p <- generic_plot( v ="cov", name = "Coverage rate") + 
+# p <- simulation_plot( v ="cov", name = "Coverage rate") + 
 #   list(geom_hline(yintercept = 0.95, color = "grey")) +
@@
 
-Cov is more or less fine, except for MNAR with only 0.05% incomplete cases --\> due to bias in the estimate in combination with very narrow CI (see CIW!). However, all stable after it=3??
+Nominal coverage is quickly reached.
+After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].
 
 ```{r echo=FALSE}
-p <- generic_plot(v="CIW", name = "CI width") 
-p %>% plotly::ggplotly()
+p <- simulation_plot(v="CIW", name = "Average CI width") 
+p # %>% plotly::ggplotly()
 ```
 
-CIW clearly determined by the proportion of incomplete cases. Worst for p=0.95, then stable after it 7.
+The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached. Depending on the proportion of incomplete cases this takes up-to $it \geq 9$.
 
 ```{r echo=FALSE}
-p <- generic_plot(v="rsq", name = "Bias in coefficient of determination") +
+p <- simulation_plot(v="rsq", name = "Bias in coefficient of determination") +
   list(geom_hline(yintercept = 0, color = "grey"))
-p %>% plotly::ggplotly()
+p # %>% plotly::ggplotly()
 ```
+Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations.
+We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
 
-Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=5.
+<!-- Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=6. -->
 
@@
 
-- need for multivar parameters!!
+-   Convergence diagnostics keep improving substantially until it = 20-30
 
--   Estimates etc do not improve after it = 7
+-   Performance measures do not improve after it = 9
 
--   Convergence diagnostics keep improving, but are all stable after 20-30 it
+-  [methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
 
--   Univariate thetas do not differentiate between missingness mechanisms or proportions of incomplete cases
+-   Univariate thetas may under-estimate non-convergence.
 
@@
 
-methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate conv slower than the est
-
-calc rhat and cov for each block of 5 it
-
-disc: conf val is likely to happen much sooner
-
-as soon as we start down-weighting the first few it from the calc, the memory effect would
+- Idea: calculate rhat for each block of 5 it
 

### 2020-10-19 Manuscript.Rmd
Add description case study plots

@@
 output:
+  html_document: default
+  pdf_document: default
   word_document: default
-  pdf_document: default
-  html_document: default
 bibliography: ref.bib
@@
 load("Data/results_25_75.Rdata")
+# load data
+load("Data/example.Rdata")
+load("Data/example_mids.Rdata")
 
@@
 }
+
+# plot each var of example data
+casestudy_plot <- function(v, name, ...){
+ggplot(example, aes(x = .data[["it"]], y = .data[[v]])) +
+  geom_line(na.rm = TRUE) +
+    labs(x = "Iteration number",
+         y = name) +
+  theme_classic()
+}
+
 ```
@@
 
+With this study, we show that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
+Any further iterations would then burn computational resources without improving the statistical inferences.
+
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the non-convergence identifiers.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
+
 -   Convergence diagnostics keep improving substantially until it = 20-30
@@
 
-## Example data
+## Case Study
 
-```{r}
-# load data
-load("Data/example.Rdata")
-load("Data/example_mids.Rdata")
+- We use real data: the `boys` dataset from the `mice` package
 
-# plot each var of example data
-plotfun <- function(v, ...){
-ggplot(example, aes(x = .data[["it"]], y = .data[[v]])) +
-  geom_line() +
-  theme_classic()
-}
+- We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`
 
-plotfun(v="est")
-plotfun(v="ac.max.qhat") #+ list(scale_y_continuous(limits = c(0.65,1)))
-plotfun(v="ac.max.lambda")
-plotfun(v="r.hat.max.sigma.hgt")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
-plotfun(v="r.hat.max.qhat")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
-plotfun(v="r.hat.max.lambda")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda
 
-plot(mids)
+```{r echo=FALSE}
+# autocorrelation plots (not used)
+# casestudy_plot(v="ac.max.qhat") #+ list(scale_y_continuous(limits = c(0.65,1)))
+# casestudy_plot(v="ac.max.lambda")
+```
+
+```{r echo=FALSE}
+p <- plot(mids)
+p[1:2,1] # %>% plotly::ggplotly()
+```
+
+```{r echo=FALSE}
+# p <- casestudy_plot(v="est", name = "Regression estimate")
+# p # %>% plotly::ggplotly()
+```
+
+```{r echo=FALSE}
+p <- casestudy_plot(v="r.hat.max.sigma.hgt", name = "Rhat chain variance")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+p # %>% plotly::ggplotly()
+```
+
+```{r echo=FALSE}
+p <- casestudy_plot(v="r.hat.max.qhat", name = "Rhat regression estimate")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+p # %>% plotly::ggplotly()
+```
+
+```{r echo=FALSE}
+p <- casestudy_plot(v="r.hat.max.lambda", name = "Rhat lambda")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+p # %>% plotly::ggplotly()
 ```

### 2020-10-20 Manuscript.Rmd
Add iterations to case study

@@
 While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
-Our study provides insight into identifying non-convergence in iterative imputation algorithms since there is no consensus on how to evaluate the convergence properties currently.
+Since there is no consensus on how to evaluate the convergence properties iterative imputation algorithms, our study provides insight into identifying non-convergence.
 We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
@@
 A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset.
-Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances (Rubin, 1987; van Buuren, 2018).
+Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
 To obtain imputations, most imputation software packages rely on iterative algorithms.
@@
 These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2).
-While there is no scientific consensus on how to evaluate the convergence of imputation algorithms (Zhu & Raghunathan, 2015; Takahashi, 2017), the current practice is to visually inspect imputations for signs of non-convergence.
+While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence.
+<!-- (Zhu & Raghunathan, 2015; Takahashi, 2017) -->
 
@@
 
-We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.0.2; R Core Team 2020).
+We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.0.3; R Core Team 2020).
 We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
@@
 
-We aim to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice].
+With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice].
 Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
-To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$).
+To induce non-convergence, we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$).
 We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].
@@
 
-We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.
+We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter that we propose, $\lambda$.
 
@@
 
-The following plots provide an overview of the results for each method and each performance measure, contrasted to the number of iteration in the imputation algorithm.
-Within the plots, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
+The following figures provide an overview of the results for each method and each performance measure, contrasted to the number of iterations in the imputation algorithm.
+Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
 
@@
 
-In this plot, our diagnostic method is the amount of autocorrelation in the chain means.
-We can interpret this as the extent to which the average imputed value trends across iterations.
-We see that autocorrelation in the chain means rapidly decreases until $it \geq 6$.
+In this figure, we see autocorrelation in the chain means as the diagnostic method.
+We can interpret this as the extent to which the average imputed value trends across iterations [unnecessary??].
+We observe that autocorrelation in the chain means rapidly decreases until $it \geq 6$.
 This means that there is some initial trending within chains, but stationarity does not improve substantively after the first few iterations.
@@
 
-Autocorrelation in the chain variances shows something similar.
-The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally when $it \geq 10$.
-We do not observe a systematic effect of the missingness conditions here either.
+Autocorrelation in the chain variances show us something similar.
+The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $it \geq 10$.
+We do not observe a systematic difference between missingness conditions here either.
 
@@
 Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$.
-There is no clear effect of the missingness mechanisms.
+There is no clear effect of the missingness mechanisms [unnecessary?].
 
@@
 
-We observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations.
-An exception to this observation is the initial increase when $3 < it < 5$.
-According to these results, the mixing between chain means generally improves until $it \geq 30$ to $40$.
+We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations.
+An exception to this observation is the initial increase when $3 \leq it \leq 5$ [interpret??
+due to initialization or is there really more mixing initially??].
+After the first couple of iterations, the mixing between chain means generally improves until $it \geq 30$ to $40$.
 There is no apparent differentiation between the missingness conditions.
@@
 With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again.
-We do, however, see some differentiation between missingness conditions.
+We do, however, see some differences between missingness conditions.
 Conditions where 75% of the cases are incomplete show more extreme non-mixing.
@@
 
-$\widehat{R}$-values in the chains of the novel parameter show a trend similar to the chain means and chain variances. [add something about conditions??]
+$\widehat{R}$-values of the novel parameter show a trend similar to the chain means and chain variances. [add something about conditions??]
 
@@
 ```
+
 Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations.
@@
 
+From the traceplot of the chain means it seems that mixing improves up-to 20 iterations, and there is some trending up-to 20 iterations as well.
+The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations.
+
 ```{r echo=FALSE}
@@
 ```{r echo=FALSE}
-p <- casestudy_plot(v="r.hat.max.sigma.hgt", name = "Rhat chain variance")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+p <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
+  list(
+    geom_hline(yintercept = 1.2, color = "grey")#, 
+    #scale_y_continuous(limits = c(1,5))
+    )
 p # %>% plotly::ggplotly()
@@
 ```{r echo=FALSE}
-p <- casestudy_plot(v="r.hat.max.qhat", name = "Rhat regression estimate")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+p <-
+  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
+  list(geom_hline(yintercept = 1.2, color = "grey")#,
+       #scale_y_continuous(limits = c(1, 2.5))
+       )
 p # %>% plotly::ggplotly()
@@
 ```{r echo=FALSE}
-p <- casestudy_plot(v="r.hat.max.lambda", name = "Rhat lambda")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
+p <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
+  list(geom_hline(yintercept = 1.2, color = "grey")#,
+       #scale_y_continuous(limits = c(1, 2.5))
+       )
 p # %>% plotly::ggplotly()

### 2020-10-20 CreateExampleR.R

@@ -2,7 +2,7 @@
 
 # setup
 source("~/MissingThePoint/R/setup_example.R")
-setup_example(seed=11)
+setup_example(seed=128)
 ests <- purrr::map_dfr(1:n_it, function(it){impute_once(iteration = it)})
 param <- preprocess_thetas(mids$chainMean, ext = "mu") %>% 
   dplyr::full_join(

### 2020-10-20 R/setup_example.R

@@ -8,7 +8,7 @@ setup_example <- function(seed){
   
   # parameters
   n_imp <<- 5
-  n_it <<- 25
+  n_it <<- 30
   mids <<- qhats <<- lambdas <<- NULL  
 }
 

### 2020-10-20 Manuscript.Rmd
Add even more iterations to case study and interpret

@@
 
-- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda
+- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.
+
+- The figures show results of a `mice` run with 30 iterations but otherwise default settings.
 
@@
 # p <- casestudy_plot(v="est", name = "Regression estimate")
-# p # %>% plotly::ggplotly()
+# p %>% plotly::ggplotly()
 ```
@@
 
+This figure shows that 36 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
+
 ```{r echo=FALSE}
@@
 
+The scientific estimate reaches the threshold much sooner, when $it=15$.
+
 ```{r echo=FALSE}
@@
 
+According to the $\widehat{R}$-values of the novel parameter, at least 25 iterations are required.
+
+[Is this a nice example? Why is convergence so slow? Don't the defaults work? Is this the bmi+wgt+wgt problem? Maybe find an easier problem?]
+
 # References (incomplete)

### 2020-10-20 R/setup_example.R

@@ -8,7 +8,7 @@ setup_example <- function(seed){
   
   # parameters
   n_imp <<- 5
-  n_it <<- 30
+  n_it <<- 50
   mids <<- qhats <<- lambdas <<- NULL  
 }
 

### 2020-10-22 Manuscript.Rmd
Re-run sim with new R version, fix non-convergence in case study

@@
 load("Data/example.Rdata")
-load("Data/example_mids.Rdata")
+load("Data/example_chains.Rdata")
 
 # sim results plotting function
-simulation_plot <- function(v, name, ...){
-  results %>%
+simulation_plot <- function(d = results, v, name, ...){
+  d %>%
     mutate(
@@
 # plot each var of example data
-casestudy_plot <- function(v, name, ...){
-ggplot(example, aes(x = .data[["it"]], y = .data[[v]])) +
+casestudy_plot <- function(d = example, v, name, ...){
+ggplot(d, aes(x = .data[["it"]], y = .data[[v]])) +
   geom_line(na.rm = TRUE) +
@@
 These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2).
-While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence.
+While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@liu14; @zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence.
 <!-- (Zhu & Raghunathan, 2015; Takahashi, 2017) -->
@@
 
--   What does non-convergence look like? --\> non-stationarity + non-mixing in a certain parameter, e.g., chain means.
+- With MCMC, the generated values will vary even after convergence, so there is not a unique point at which convergence is established [@gelm13].
+Therefore, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
+Non-convergence occurs when one of two requirements for convergence is not met.
 
--   What are the consequences? --\> bias, under-coverage --\> refer to van Buuren (2018) instead of reproducing this
+- There are two requirements for convergence of iterative algorithms: mixing and
+stationarity [@gelm13]. Mixing implies that generated values intermingle nicely, and stationarity is characterized by the absence of trending between successive draws.
+
+-   What are the consequences? --\> bias, under-coverage [refer to van Buuren (2018) instead of reproducing this]
 
@@
 
-The following figures provide an overview of the results for each method and each performance measure, contrasted to the number of iterations in the imputation algorithm.
+The following figures display the simulation results for each method and each performance measure, contrasted to the number of iterations in the imputation algorithm.
 Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
@@
 
-In this figure, we see autocorrelation in the chain means as the diagnostic method.
-We can interpret this as the extent to which the average imputed value trends across iterations [unnecessary??].
 We observe that autocorrelation in the chain means rapidly decreases until $it \geq 6$.
-This means that there is some initial trending within chains, but stationarity does not improve substantively after the first few iterations.
+This means that there is some initial trending within chains, but stationarity of the average imputed value does not improve substantively after the first few iterations.
 These results hold irrespective of the missingness condition.
@@
 ```{r echo=FALSE}
-p <- plot(mids)
-p[1:2,1] # %>% plotly::ggplotly()
+# p <- plot(mids)
+# p[1:2,1] # %>% plotly::ggplotly()
+param %>% 
+  ggplot() + 
+  geom_line(aes(x = it, y = mu.hgt, color = as.factor(m))) + 
+  theme_classic() + 
+  labs(x = "Iteration number",
+       y = "Chain mean variable 'hgt'",
+       color = "Imputation")
 ```
@@
 
-This figure shows that 36 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
+This figure shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
 
@@
 
-The scientific estimate reaches the threshold much sooner, when $it=15$.
+The scientific estimate reaches the threshold much sooner, when $it=14$.
 
@@
 
-According to the $\widehat{R}$-values of the novel parameter, at least 25 iterations are required.
+According to the $\widehat{R}$-values of the novel parameter, at least 15 iterations are required.
 

### 2020-10-22 CreateExampleR.R

@@ -2,15 +2,16 @@
 
 # setup
 source("~/MissingThePoint/R/setup_example.R")
-setup_example(seed=128)
+setup_example(seed=1)
 ests <- purrr::map_dfr(1:n_it, function(it){impute_once(iteration = it)})
 param <- preprocess_thetas(mids$chainMean, ext = "mu") %>% 
   dplyr::full_join(
     preprocess_thetas(mids$chainVar, ext = "sigma") %>% 
       dplyr::full_join(qhats) %>% 
-      dplyr::full_join(lambdas)
+      dplyr::full_join(lambdas) %>% 
+      dplyr::full_join(svs)
   )
 convs <- get_diagnostics(param)
 example <- full_join(ests, convs)
 save(example, file = "Data/example.Rdata")
-save(mids, file = "Data/example_mids.Rdata")
+save(param, file = "Data/example_chains.Rdata")

### 2020-10-22 R/impute.R

@@ -47,6 +47,7 @@ impute <-
     lambda <- mild %>%
       purrr::map_dbl(., ~ {
         princomp(., cor = TRUE) %>% .$sdev %>% .[1] %>% . ^ 2 #first eigenvalue of the varcovar matrix
+        # svd(.) %>% .$d %>% .[1]
       })
     
     # save for each simulation condition

### 2020-10-22 R/setup_example.R

@@ -8,8 +8,8 @@ setup_example <- function(seed){
   
   # parameters
   n_imp <<- 5
-  n_it <<- 50
-  mids <<- qhats <<- lambdas <<- NULL  
+  n_it <<- 20
+  mids <<- qhats <<- lambdas <<- svs <<- NULL  
 }
 
 # other functions to load
@@ -49,18 +49,29 @@ get_diagnostics <- function(parameters) {
 }
 
 impute_once <- function(iteration, ...) {
+  # specify correct imputation model and predictor matrix
+  meth <- make.method(boys)
+  meth["bmi"] <- "~ I(wgt / (hgt / 100)^2)"
+  pred <- make.predictorMatrix(boys)
+  pred[c("hgt", "wgt"), "bmi"] <- 0
+  # impute once to create mids object
   if(is.null(mids)){
-    mids <<- mice(boys, m = n_imp, maxit = 1, printFlag = FALSE)
+    mids <<- mice(boys, 
+                  meth = meth, 
+                  pred = pred, 
+                  m = n_imp, 
+                  maxit = 1, 
+                  printFlag = FALSE)
   }
   else {
     mids <<- mice.mids(mids, printFlag = FALSE)
   }
+  # extract estimates and diagnostics
   mira <- mids %>% mice::lm.mids(age ~ ., data = .) 
   mipo <- mira %>%
     mice::pool() %>%
     summary(., conf.int = TRUE)
   mild <- mids %>% mice::complete("all")
-  
   qhat <- 
     map_dbl(mild, ~ {
       lm(formula = age ~ ., data = .) %>% .$coefficients %>% .[2]
@@ -68,9 +79,16 @@ impute_once <- function(iteration, ...) {
   lambda <- mild %>%
     purrr::map_dbl(., ~ {
       dplyr::mutate(., dplyr::across(dplyr::everything(), ~as.numeric(.))) %>% 
+        # svd(.) %>% .$d %>% .[1]
         princomp(., cor = TRUE) %>% .$sdev %>% .[1] %>% . ^ 2 #first eigenvalue of the varcovar matrix
     })
+  sv <- mild %>%
+    purrr::map_dbl(., ~ {
+      dplyr::mutate(., dplyr::across(dplyr::everything(), ~as.numeric(.))) %>% 
+        svd(.) %>% .$d %>% .[1]
+    })
[…]

### 2020-10-22 RunSimulation.R

@@ -8,7 +8,7 @@ source("R/preprocess.R")
 source("R/ComputeDiagnostics.R")
 setup(seed = 1111)
 it_total <- 50
-n_sim <- 100
+n_sim <- 1000
 
 # create simulation function
 simulate <- function(m_mech, p_inc, amp_pat, it_total, n_imp, ...){

### 2020-11-13 Manuscript.Rmd
Comments Gerko 12/11

@@
 Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences.
-A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset.
-Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
-To obtain imputations, most imputation software packages rely on iterative algorithms.
+The de facto standard to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset.
+To generate imputed values, most imputation software packages rely on iterative algorithms.
+[Why iterative?
+We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.] Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
 
@@
 And can we rely on the inference obtained using the imputed data?
-These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2).
-While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@liu14; @zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence.
-<!-- (Zhu & Raghunathan, 2015; Takahashi, 2017) -->
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+[Common convergence diagnostics may not work.] While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@liu14; @zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence.
+<!-- (Van Buuren, 2018, 6.5.2) (Zhu & Raghunathan, 2015; Takahashi, 2017) -->
 
@@
 
-In this paper, we explore diagnostic methods for iterative imputation algorithms.
+In this paper, we explore some existing and new means of diagnosing non-convergence in iterative algorithms.
+We evaluate the validity of these methods in the context of missing data imputation using model-based simulation in `R` [@R].
+[Brief overview of paper: the diagnostics, the simulation, the results, the discussion, a case study.]
+
+<!-- In this paper, we explore diagnostic methods for iterative imputation algorithms.
 For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
@@
 Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm.
+-->
 
-We propose a novel multivariate parameter to check for non-convergence in iterative algorithms.
+<!-- We propose a novel multivariate parameter to check for non-convergence in iterative algorithms.
 We aim to show which method is the most informative about non-convergence in iterative imputation [explain that method = parameter + diagnostic + interpretation].
+-->
 
@@
 
-    for each simulation repetition (1:1000)
+    for each simulation repetition 
       1. simulate complete data
-      for each missingness condition (1:9)
+      for each missingness condition 
         2. create missingness
-        for each iteration (1:50)
+        for each iteration 
           3. impute missingness
@@
 
-With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice].
+With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice].
 Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
@@
 
-[Add mvtnorm package here??] The complete set is amputed according to nine missingness conditions.
+The complete set is amputed according to nine missingness conditions.
 We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.
@@
 $$
-Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
+\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
 $$
-where $Y'$ is the expected value of the outcome.
-Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $r^2$ that we obtain after pooling the regression results across the imputations.
+where $\hat{Y}$ is the predicted value of the outcome.
+Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $\rho^2$ that we obtain after pooling the regression results across the imputations.
 
@@
 
-As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($E(\bar{Q}_{LL} - \bar{Q}_{UL})$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands [or just of the regression estimate??
-otherwise add the ciw and cov of r2 too!!
-and check if definition of ciw is correct].
+As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2)[or just of the regression estimate??
+otherwise add the ciw and cov of r2 too!!].
 
@@
 
-The following figures display the simulation results for each method and each performance measure, contrasted to the number of iterations in the imputation algorithm.
+The following figures display the simulation results for the eight diagnostic methods and four performance measures, contrasted to the number of iterations in the imputation algorithm.
 Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
+Note that these results are averages of the $n_{sim} = 1000$ simulation repetitions.
 
@@
 
-We observe that autocorrelation in the chain means rapidly decreases until $it \geq 6$.
-This means that there is some initial trending within chains, but stationarity of the average imputed value does not improve substantively after the first few iterations.
+Autocorrelation in the chain means decreases rapidly in the first few iterations.
+The decrease is substantive until $it \geq 6$.
+This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity.
 These results hold irrespective of the missingness condition.
@@
 
-With this study, we show that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
-Any further iterations would then burn computational resources without improving the statistical inferences.
+The validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought.
+We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
+Any further iterations just burn computational resources without improving the statistical inferences.
 
@@
 
-- The figures show results of a `mice` run with 30 iterations but otherwise default settings.
+- The figures show results of a `mice` run with 20 iterations but otherwise default settings.
 
@@
 
-From the traceplot of the chain means it seems that mixing improves up-to 20 iterations, and there is some trending up-to 20 iterations as well.
-The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations.
+From the traceplot of the chain means it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
+
+<!-- and there is some trending up-to 20 iterations as well. The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations. -->
 
@@
 
-[Is this a nice example? Why is convergence so slow? Don't the defaults work? Is this the bmi+wgt+wgt problem? Maybe find an easier problem?]
-
 # References (incomplete)

### 2020-11-13 Manuscript.Rmd
Combine figures and add labels

@@
 ---
-title: "Missing The Point"
+title: "Missing The Point: Non-Convergence in Iterative Imputation Algorithms"
 author: "Hanne Oberman"
@@
 output:
-  html_document: default
-  pdf_document: default
-  word_document: default
+  bookdown::html_document2
 bibliography: ref.bib
@@
 # environment
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.width = 10)
 library(tidyverse)
 library(mice)
+library(patchwork)
+# load data
+load("Data/parameters.Rdata")
 load("Data/results_25_75.Rdata")
-# load data
 load("Data/example.Rdata")
@@
 
-# Missing The Point: Non-Convergence in Iterative Imputation Algorithms
-
 *Notation in this manuscript: *
@@
 
-## Abstract
+# Abstract
 
@@
 
-## Intro
+# Intro
 
@@
 
-## Identifying non-convergence
+# Identifying non-convergence
 
@@
 
-## Simulation Set-Up
+```{r eval=FALSE, include=FALSE}
+# # maybe add ESS instead of AC???
+# LaplacesDemon::ESS(parameters[parameters$m == 1 & parameters$sim == 1, "qhat"])
+```
+
+# Simulation Set-Up
 
@@
 
-### Aims
+## Aims
 
@@
 
-### Data generating mechanism
+## Data generating mechanism
 
@@
 
-### Estimands
+## Estimands
 
@@
 
-### Methods
+## Methods
 
@@
 
-### Performance measures
+## Performance measures
 
@@
 
-## Simulation Results
+# Simulation Results
 
@@
 
-### Diagnostic Methods
+## Diagnostic Methods
 
+Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
 <!-- The next four plots show the autocorrelations in each parameter plotted against the number of iterations.  -->
 
-```{r echo=FALSE}
-p <- simulation_plot(v="ac.max.mu.Y", name = "Autocorrelation chain means")
-p # %>% plotly::ggplotly()
+```{r ac, fig.cap = "Autocorrelation (AC) with different parameters.", echo=FALSE}
+ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") 
+ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances")
+ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta))
+ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda))
+(ac_means + ac_vars) / (ac_betas + ac_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
+
 ```
 
-Autocorrelation in the chain means decreases rapidly in the first few iterations.
+Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A).
 The decrease is substantive until $it \geq 6$.
@@
 
-```{r echo=FALSE}
-p <- simulation_plot(v="ac.max.sigma.Y", name = "Autocorrelation chain variances")
-p # %>% plotly::ggplotly() 
-```
-
-Autocorrelation in the chain variances show us something similar.
+Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B).
 The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $it \geq 10$.
@@
 
-```{r echo=FALSE}
-p <- simulation_plot(v="ac.max.qhat", name = "Autocorrelation scientific estimate")
-p # %>% plotly::ggplotly() 
-```
-
-There is more autocorrelation in the scientific estimates than in the univariate parameters (chain means and chain variances).
+There is more autocorrelation in the scientific estimates than in the chain means and chain variances (see \@ref(fig:ac)C).
 We observe the highest autocorrelations in conditions where 75% of cases are incomplete.
@@
 
-```{r echo=FALSE}
-p <- simulation_plot(v="ac.max.lambda", name = "Autocorrelation novel theta")
-p # %>% plotly::ggplotly() 
+The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $it \geq 20$.
+
+Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
+<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->
+```{r rh, fig.cap = "Potential scale reduction factor (Rh) with different parameters.", echo=FALSE}
+rh_means <- simulation_plot(v="r.hat.max.mu.Y", name = "Rh chain means") 
+rh_vars <- simulation_plot(v="r.hat.max.sigma.Y", name = "Rh chain variances")
+rh_betas <- simulation_plot(v="r.hat.max.qhat", name = quote("Rh"~beta))
+rh_lambdas <- simulation_plot(v="r.hat.max.lambda", name = quote("Rh"~lambda))
+(rh_means + rh_vars) / (rh_betas + rh_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
+
 ```
 
-The autocorrelation in the novel parameter exhibits a similar trend to the autocorrelation in the scientific estimates. Trending in this parameter diminishes when $it \geq 20$.
-
-<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->
-
-```{r echo=FALSE}
-p <- simulation_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
-p # %>% plotly::ggplotly() 
-```
-
-We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations.
+We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations (see \@ref(fig:rh)A).
 An exception to this observation is the initial increase when $3 \leq it \leq 5$ [interpret??
@@
 
-```{r echo=FALSE}
-p <- simulation_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
-p # %>% plotly::ggplotly() 
-```
+The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B).
+Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $it \geq 30$.
 
-The mixing between chain variances mimics the mixing between chain means almost perfectly. Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $it \geq 30$.
-
-```{r echo=FALSE}
-p <- simulation_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
-p # %>% plotly::ggplotly() 
-```
-
-With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again.
+With the regression estimate as parameter we observe very similar $\widehat{R}$-values than with chain means and chain variances (see \@ref(fig:rh)C).
 We do, however, see some differences between missingness conditions.
@@
 
-```{r echo=FALSE}
-p <- simulation_plot(v="r.hat.max.lambda", name = "R hat novel theta")
-p # %>% plotly::ggplotly() 
-```
-
-$\widehat{R}$-values of the novel parameter show a trend similar to the chain means and chain variances. [add something about conditions??]
+$\widehat{R}$-values of the eigenvalues show a trend similar to the chain means and chain variances (see \@ref(fig:rh)D). [add something about conditions??]
 
@@
 
-### Performance Measures
+## Performance Measures
 
@@
 
-## Discussion
+# Discussion
 
@@
 
-<!-- ## Check if heuristic for AC works -->
+<!-- # Check if heuristic for AC works -->
 
@@
 
-## Case Study
+# Case Study
 

### 2020-11-18 Manuscript.Rmd
Combine performance measures plots, add feedback Gerko

@@
 
-Iterative imputation is a popular tool to accommodate the ubiquitous problem of missing data.
+Iterative imputation has become the de facto approach to accommodate the ubiquitous problem of missing data.
 While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
-Since there is no consensus on how to evaluate the convergence properties iterative imputation algorithms, our study provides insight into identifying non-convergence.
+Since there is no consensus on how to evaluate the convergence properties of iterative imputation algorithms, our study provides insight into identifying non-convergence.
 We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
@@
 ```{r ac, fig.cap = "Autocorrelation (AC) with different parameters.", echo=FALSE}
-ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") 
-ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances")
-ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta))
-ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda))
+ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") + ggplot2::scale_y_continuous(limits = c(0.15, 1))
+ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances") + ggplot2::scale_y_continuous(limits = c(0.15, 1))
+ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta)) + ggplot2::scale_y_continuous(limits = c(0.15, 1))
+ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda)) + ggplot2::scale_y_continuous(limits = c(0.15, 1))
 (ac_means + ac_vars) / (ac_betas + ac_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
@@
 ```{r rh, fig.cap = "Potential scale reduction factor (Rh) with different parameters.", echo=FALSE}
-rh_means <- simulation_plot(v="r.hat.max.mu.Y", name = "Rh chain means") 
-rh_vars <- simulation_plot(v="r.hat.max.sigma.Y", name = "Rh chain variances")
-rh_betas <- simulation_plot(v="r.hat.max.qhat", name = quote("Rh"~beta))
-rh_lambdas <- simulation_plot(v="r.hat.max.lambda", name = quote("Rh"~lambda))
+rh_means <- simulation_plot(v="r.hat.max.mu.Y", name = "Rh chain means") + ggplot2::scale_y_continuous(limits = c(1, 1.65))
+rh_vars <- simulation_plot(v="r.hat.max.sigma.Y", name = "Rh chain variances") + ggplot2::scale_y_continuous(limits = c(1, 1.65))
+rh_betas <- simulation_plot(v="r.hat.max.qhat", name = quote("Rh"~beta)) + ggplot2::scale_y_continuous(limits = c(1, 1.65))
+rh_lambdas <- simulation_plot(v="r.hat.max.lambda", name = quote("Rh"~lambda)) + ggplot2::scale_y_continuous(limits = c(1, 1.65))
 (rh_means + rh_vars) / (rh_betas + rh_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
@@
 
-```{r echo=FALSE}
-p <- simulation_plot(v = "est", name = "Bias in regression estimate") + 
+```{r perf, fig.cap = "Performance measures.", echo=FALSE}
+est <- simulation_plot(v = "est", name = "Bias in regression estimate") + 
   list(geom_hline(yintercept = 0, color = "grey")) 
-p # %>% plotly::ggplotly()
+cov <- simulation_plot( v ="cov", name = "Coverage rate") + 
+  list(geom_hline(yintercept = 0.95, color = "grey"))
+ciw <- simulation_plot(v="CIW", name = "Average CI width") 
+rsq <- simulation_plot(v="rsq", name = "Bias in coefficient of determination") +
+  list(geom_hline(yintercept = 0, color = "grey"))
+(est + rsq) / (cov + ciw) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
 ```
 
-We see that within a few iterations the average bias approaches zero.
+We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A).
 When $it \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
+
+Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations (see \@ref(fig:perf)B).
+We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
 
@@
 ```{r echo=FALSE}
-p <- simulation_plot( v ="cov", name = "Coverage rate") + 
-  list(geom_hline(yintercept = 0.95, color = "grey"))
-p # %>% plotly::ggplotly()
-
 # p <- simulation_plot( v ="cov", name = "Coverage rate") + 
@@
 
-Nominal coverage is quickly reached.
+Nominal coverage is quickly reached (see \@ref(fig:perf)C).
 After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].
 
-```{r echo=FALSE}
-p <- simulation_plot(v="CIW", name = "Average CI width") 
-p # %>% plotly::ggplotly()
-```
-
-The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached. Depending on the proportion of incomplete cases this takes up-to $it \geq 9$.
-
-```{r echo=FALSE}
-p <- simulation_plot(v="rsq", name = "Bias in coefficient of determination") +
-  list(geom_hline(yintercept = 0, color = "grey"))
-p # %>% plotly::ggplotly()
-```
-
-Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations.
-We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
+The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D).
+Depending on the proportion of incomplete cases this takes up-to $it \geq 9$.
 

### 2020-11-18 Manuscript.Rmd
Add reference to intro

@@
 
-Iterative imputation has become the de facto approach to accommodate the ubiquitous problem of missing data.
+Iterative imputation has become the de facto standard approach to accommodate the ubiquitous problem of missing data.
 While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
-Since there is no consensus on how to evaluate the convergence properties of iterative imputation algorithms, our study provides insight into identifying non-convergence.
-We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
+Our study provides insight into identifying non-convergence in iterative imputation algorithms.
+We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
+In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
 We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
@@
 
-Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences.
-The de facto standard to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset.
+<!-- Iterative imputation has become the de facto standard approach to accommodate missing data problems.
+With iterative imputation, we 'impute' (i.e., fill in) plausible values in an incomplete dataset.
+These plausible values are generated using algorithms such as 'MICE' [@mice].
+These algorithms sequence through the variables in an incomplete dataset  -->
+
+<!-- Iterative imputation entails using algorithms such as MICE [@mice] to replace missing values by plausible values.
+These plausible values are sampled from the distribution of each missing entry, and draw a plausible value from this distribution to replace the missing value by.
+-->
+
+<!-- While it is widely accepted that the technique can yield valid inferences under many different missing data circumstances, these inferences all rely on algorithmic convergence.
+We fill in the missing data by modeling a multivariate distribution for each missing entry.
+We update dis model by sequencing through the variables See Figure 1.
+Van Buuren et al.
+(2006) have shown that just a few iterations maybe sufficient.
+Most imputation software packages use a default value of five or ten iterations.
+But how can we know did this is sufficient?
+-->
+<!-- There is currently no consensus on how to evaluate the convergence properties of iterative imputation algorithms.
+Imputation algorithms such as mice are a type of Gibbs sampler.
+It would does make sense to evaluate the convergence of iterative imputation algorithms using common MSMC non convergence identifiers.
+This has not been done.
+-->
+<!-- With MCMC algorithms there is not a single point at which convergence is reached.  -->
+
+Iterative imputation has become the de facto standard approach to accommodate missing data.
+With iterative imputation, we 'impute' (i.e., fill in) plausible data values in the place of missing values in an incomplete dataset.
+We obtain these plausible values by sampling from a distribution that we model for each missing entry and update sequentially.
+@buur06 have shown that with an iterative imputation algorithm such as 'MICE', just a few iterations are enough to yield unbiased, confidence-valid statistical inferences.
+
+<!-- Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences.
+The de facto approach to accommodate missing data is to 'impute' (i.e., fill in) any missing entries in an incomplete data.
+With iterative imputation, we model the distribution of each missing entry, and draw one or more plausible values from it.
+-->
+
+<!-- With iterative imputation, we 'impute' (i.e., fill in) the missing data in an incomplete dataset by drawing values from the distribution of each missing entry.
 To generate imputed values, most imputation software packages rely on iterative algorithms.
 [Why iterative?
-We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.] Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
+We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.]  -->
+
+Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
 

### 2020-11-18 Manuscript.Rmd
Add new bullet point intro, combine figures in case study part

@@
 
+## One
+
+- iterative imputation has become the de facto standard approach to accommodate missing data
+
+- the aim is to draw valid inference: to get unbiased, confidence-valid estimates
+
+- the estimates are obtained by separating the missing data problem from the scientific problem
+
+- the missing data problem is solved by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE
+
+- then the analysis of scientific interest can be performed on the completed data, which yields the scientific estimates
+
+- to obtain valid estimates, both the missing data problem and the scientific problem should be solved correctly
+
+- one aspect that is often overlooked is the convergence of the algorithm that is used to generate imputations
+
+## Two
+
+- there is no consensus on the convergence properties of iterative imputation, and there has not been a systematic study on how to evaluate the convergence of these algorithms
+
+- the current practice is visual inspection, but this may not be adequate because 1, 2, 3 (see original text)
+
+- it is difficult because there is not at single point at which convergence is reached, because the algorithm will produce some fluctuation even after convergence (and a ...
+distribution may not exist because of univariate modeling of multivariate distribution???)
+
+- therefore, we should look at *non*-convergence
+
+- there are non-convergence identifiers for other iterative algorithms (such as MCMC), but it is not known whether these are appropriate for imputation algorithms as well
+
+## Three
+
+- in this paper, we explore different methods for identifying non-convergence in iterative imputation algorithms
+
+- since the ultimate goal is to have inferential validity, we use this as performance measure in a simulation study: if the estimates are unbiased and confidence-valid, the non-convergence is not influential, apparently
+
+- since we typically don't have the ground truth (to calculate bias and coverage), we develop guidelines for interpreting the non-convergence identifiers in practice
+
 <!-- Iterative imputation has become the de facto standard approach to accommodate missing data problems.
@@
 <!-- With MCMC algorithms there is not a single point at which convergence is reached.  -->
+
+## Original text
 
@@
 
-```{r echo=FALSE}
-# p <- simulation_plot( v ="cov", name = "Coverage rate") + 
-#   list(geom_hline(yintercept = 0.95, color = "grey")) +
-#   ylim(0.85, 1)
-
-```
-
 Nominal coverage is quickly reached (see \@ref(fig:perf)C).
@@
 
-```{r echo=FALSE}
-# autocorrelation plots (not used)
-# casestudy_plot(v="ac.max.qhat") #+ list(scale_y_continuous(limits = c(0.65,1)))
-# casestudy_plot(v="ac.max.lambda")
-```
-
-```{r echo=FALSE}
-# p <- plot(mids)
-# p[1:2,1] # %>% plotly::ggplotly()
-param %>% 
+```{r case, fig.cap = "Case study.", echo=FALSE}
+trace <- param %>% 
   ggplot() + 
@@
        color = "Imputation")
+means <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
+  list(geom_hline(yintercept = 1.2, color = "grey"))
+qhats <-
+  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
+  list(geom_hline(yintercept = 1.2, color = "grey"))
+lambdas <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
+  list(geom_hline(yintercept = 1.2, color = "grey"))
+(trace + means) / (qhats + lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
+
 ```
 
-From the traceplot of the chain means it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
+From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
 
-<!-- and there is some trending up-to 20 iterations as well. The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations. -->
+This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
 
-```{r echo=FALSE}
-# p <- casestudy_plot(v="est", name = "Regression estimate")
-# p %>% plotly::ggplotly()
-```
+The scientific estimate reaches the threshold much sooner, when $it=14$ (see \@ref(fig:case)C).
 
-```{r echo=FALSE}
-p <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
-  list(
-    geom_hline(yintercept = 1.2, color = "grey")#, 
-    #scale_y_continuous(limits = c(1,5))
-    )
-p # %>% plotly::ggplotly()
-```
-
-This figure shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
-
-```{r echo=FALSE}
-p <-
-  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
-  list(geom_hline(yintercept = 1.2, color = "grey")#,
-       #scale_y_continuous(limits = c(1, 2.5))
-       )
-p # %>% plotly::ggplotly()
-```
-
-The scientific estimate reaches the threshold much sooner, when $it=14$.
-
-```{r echo=FALSE}
-p <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
-  list(geom_hline(yintercept = 1.2, color = "grey")#,
-       #scale_y_continuous(limits = c(1, 2.5))
-       )
-p # %>% plotly::ggplotly()
-```
-
-According to the $\widehat{R}$-values of the novel parameter, at least 15 iterations are required.
+According to the $\widehat{R}$-values of the novel parameter, at least 15 iterations are required (see \@ref(fig:case)D).
 

### 2020-11-19 Manuscript.Rmd
Add improved methods in bullet points

@@
 author: "Hanne Oberman"
-date: "7-9-2020"
+date: "`r format(Sys.time(), '%d %B %Y')`"
 output:
-  bookdown::html_document2
+  bookdown::pdf_document2
 bibliography: ref.bib
@@
 
-- the aim is to draw valid inference: to get unbiased, confidence-valid estimates
+- the aim is to draw valid inference: to get unbiased, confidence-valid estimates that incorporate the effect of the missingness
 
@@
 
-- the missing data problem is solved by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE
+- the missing data problem is solved first: by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE
 
@@
 
-- to obtain valid estimates, both the missing data problem and the scientific problem should be solved correctly
+- to obtain valid scientific estimates, both the missing data problem and the scientific problem should be solved correctly
 
@@
 
-- there is no consensus on the convergence properties of iterative imputation, and there has not been a systematic study on how to evaluate the convergence of these algorithms
+- there is no consensus on the convergence properties of iterative imputation algorithms, and there has not been a systematic study on how to evaluate the convergence of these algorithms
 
-- the current practice is visual inspection, but this may not be adequate because 1, 2, 3 (see original text)
+- yet, the inferences rely on algorithmic convergence
 
-- it is difficult because there is not at single point at which convergence is reached, because the algorithm will produce some fluctuation even after convergence (and a ...
-distribution may not exist because of univariate modeling of multivariate distribution???)
+- the current practice is visual inspection, but this may be inadequate because of reasons 1, 2, and 3 [see original text]
+
+- it is difficult that there is not at single point at which convergence is reached: the algorithm will produce some fluctuations even after convergence [it's convergence in distribution, not to a point]  <!-- (and a ...
+distribution may not exist because of univariate modeling of multivariate distribution???) -->
 
@@
 
-- there are non-convergence identifiers for other iterative algorithms (such as MCMC), but it is not known whether these are appropriate for imputation algorithms as well
+- there are non-convergence identifiers for other iterative algorithms (MCMC), but it is not known whether these are appropriate for imputation algorithms as well
 
@@
 
-- since the ultimate goal is to have inferential validity, we use this as performance measure in a simulation study: if the estimates are unbiased and confidence-valid, the non-convergence is not influential, apparently
+- since the ultimate goal is to have inferential validity, we use this as performance measure in a simulation study: if the estimates are unbiased and confidence-valid, the non-convergence is, apparently, not influential
 
-- since we typically don't have the ground truth (to calculate bias and coverage), we develop guidelines for interpreting the non-convergence identifiers in practice
+- we develop guidelines for practice to interpret the non-convergence identifiers, because we typically don't have the ground truth (to calculate bias and coverage)
+
+- we show how all this works by means of a case study
 
@@
 
-## Original text
+## Original text [remove later]
 
@@
 
-# Identifying non-convergence
+# Identifying non-convergence [remove later]
 
@@
 Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
-To induce non-convergence, we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$).
-We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].
+To induce non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
+We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.
+<!-- : three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??]. -->
 
@@
 
-The complete set is amputed according to nine missingness conditions.
-We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.
-[Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences).
-And add the `ampute` function?]
+The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.
+
+- we use the three missingness mechanisms MCAR, MAR, and MNAR with defaults settings from `mice::ampute()` (e.g., right-tailed MAR)
+
+- we use a multivariate missing data pattern and make 25%, 50%, and 75% of cases incomplete
 
@@
 $$
+
 where $\hat{Y}$ is the predicted value of the outcome.
@@
 
-We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter that we propose, $\lambda$.
+- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter
+
+- the parameter can be any statistic that we track across iterations, for example chain means (i.e, the average imputed value per imputation)
+
+- the identifier is a calculation of some sort that quantifies non-convergence
+
+- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains
+
+- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well
+
+- just to be sure, we also use autocorrelation to quantify trending within chains
+
+- we apply these identifiers to four different univariate and multivariate parameters
+
+- the univariate parameters are those commonly used for visual inspection: chain means and chain variances
+
+- one multivariate parameter that is of immediate interest for our estimand is the estimated value itself: regression coefficient beta_1
+
+- we also propose a new multivariate parameter that is not dependent on the model of scientific interest: the first eigenvalue of the variance-covariance matrix of the completed data [explain!]
+
+We use eight different methods to diagnose non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters---chain means, chain variances, $\beta_1$, and $\lambda$.
 
@@
 ```{r ac, fig.cap = "Autocorrelation (AC) with different parameters.", echo=FALSE}
-ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") + ggplot2::scale_y_continuous(limits = c(0.15, 1))
-ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances") + ggplot2::scale_y_continuous(limits = c(0.15, 1))
-ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta)) + ggplot2::scale_y_continuous(limits = c(0.15, 1))
-ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda)) + ggplot2::scale_y_continuous(limits = c(0.15, 1))
+ac_means <- simulation_plot(v="ac.max.mu.Y", name = "AC chain means") + ggplot2::scale_y_continuous(limits = c(0.175, 1))
+ac_vars <- simulation_plot(v="ac.max.sigma.Y", name = "AC chain variances") + ggplot2::scale_y_continuous(limits = c(0.175, 1))
+ac_betas <- simulation_plot(v="ac.max.qhat", name = quote("AC"~beta)) + ggplot2::scale_y_continuous(limits = c(0.175, 1))
+ac_lambdas <- simulation_plot(v="ac.max.lambda", name = quote("AC"~lambda)) + ggplot2::scale_y_continuous(limits = c(0.175, 1))
 (ac_means + ac_vars) / (ac_betas + ac_lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
@@
 Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A).
-The decrease is substantive until $it \geq 6$.
+The decrease is substantive until $n_{it} \geq 6$.
 This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity.
@@
 Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B).
-The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $it \geq 10$.
+The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $n_{it} \geq 10$.
 We do not observe a systematic difference between missingness conditions here either.
@@
 We observe the highest autocorrelations in conditions where 75% of cases are incomplete.
-Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$.
-There is no clear effect of the missingness mechanisms [unnecessary?].
+Overall, the autocorrelations reach a plateau when $n_{it} \geq 20$ to $30$.
+There is no clear effect of the missingness mechanisms.
 
-The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $it \geq 20$.
+The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $n_{it} \geq 20$.
 
@@
 <!-- The next four plots show the $\widehat{R}$ in the different parameters. -->
+
 ```{r rh, fig.cap = "Potential scale reduction factor (Rh) with different parameters.", echo=FALSE}
@@
 We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations (see \@ref(fig:rh)A).
-An exception to this observation is the initial increase when $3 \leq it \leq 5$ [interpret??
+An exception to this observation is the initial increase when $3 \leq n_{it} \leq 5$ [interpret??
 due to initialization or is there really more mixing initially??].
-After the first couple of iterations, the mixing between chain means generally improves until $it \geq 30$ to $40$.
+After the first couple of iterations, the mixing between chain means generally improves until $n_{it} \geq 30$ to $40$.
 There is no apparent differentiation between the missingness conditions.
@@
 The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B).
-Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $it \geq 30$.
+Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$.
 
@@
 ## Performance Measures
+
+In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), bias in the coefficient of determination (panel B), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
 
@@
 We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A).
-When $it \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
+When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
 
 Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations (see \@ref(fig:perf)B).
-We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
+We observe stable estimates in all conditions when $n_{it} \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
 
@@
 The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D).
-Depending on the proportion of incomplete cases this takes up-to $it \geq 9$.
+Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
 
@@
 
-The validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought.
-We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
-Any further iterations just burn computational resources without improving the statistical inferences.
-
 Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the non-convergence identifiers.
@@
 
--   Convergence diagnostics keep improving substantially until it = 20-30
+- Convergence diagnostics keep improving substantially until n_it = 20-30
 
--   Performance measures do not improve after it = 9
+- Performance measures do not improve after n_it = 9
 
@@
 
-- Idea: calculate rhat for each block of 5 it
+In short, the validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought.
+We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
+Any further iterations just burn computational resources without improving the statistical inferences.
 
@@
 
-The scientific estimate reaches the threshold much sooner, when $it=14$ (see \@ref(fig:case)C).
+The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).
 
-According to the $\widehat{R}$-values of the novel parameter, at least 15 iterations are required (see \@ref(fig:case)D).
+According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).
 
-# References (incomplete)
+# References

### 2020-11-26 Manuscript.Rmd
Comments gerko

@@
 
-Iterative imputation has become the de facto standard approach to accommodate the ubiquitous problem of missing data.
+Iterative imputation has become the de facto standard to accommodate for the ubiquitous problem of missing data.
 While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
@@
 
-- iterative imputation has become the de facto standard approach to accommodate missing data
+- Iterative imputation has become the de facto standard to accommodate for missing data.
 
-- the aim is to draw valid inference: to get unbiased, confidence-valid estimates that incorporate the effect of the missingness
+- the aim is usually to draw valid inferences, i.e. to get 
+<!-- ... the aim is to draw valid inference: to get  -->
+unbiased, confidence-valid estimates that incorporate the effects of the missingness
 
-- the estimates are obtained by separating the missing data problem from the scientific problem
+- such estimates are obtained in it imp by separating the missing data problem from the scientific problem
 
-- the missing data problem is solved first: by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE
+- the missing values are imputed (i.e., filled in) using some sort of algorithm. And subsequently, the scientific model of int is performed on the completed data.
 
-- then the analysis of scientific interest can be performed on the completed data, which yields the scientific estimates
+<!-- data problem is solved first: by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE -->
 
-- to obtain valid scientific estimates, both the missing data problem and the scientific problem should be solved correctly
+<!-- - then the analysis of scientific interest can be performed on the completed data, which yields the scientific estimates -->
 
-- one aspect that is often overlooked is the convergence of the algorithm that is used to generate imputations
+- to obtain valid scientific estimates, both the mis dat prob and the sci prob should be considered appropriately. 
+<!-- both the missing data problem and the scientific problem should be solved correctly -->
+
+- the validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imp. 
+<!-- one aspect that is often overlooked is the convergence of the algorithm that is used to generate imputations -->
 
@@
 
-- there is no consensus on the convergence properties of iterative imputation algorithms, and there has not been a systematic study on how to evaluate the convergence of these algorithms
+- determining whether an alg has conv is not trivial. 
+<!-- there is no consensus on the convergence properties of iterative imputation algorithms, and  -->
+there has not been a systematic study on how to evaluate the convergence of it imp algorithms
 
-- yet, the inferences rely on algorithmic convergence
+- yet, all inferences rely on the convergence of such alg
 
-- the current practice is visual inspection, but this may be inadequate because of reasons 1, 2, and 3 [see original text]
+- a widely accepted practice is visual inspection, but identifying convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence (Van Buuren, 2018, 6.5.2).
+Therefore, a quantitative diagnostic method to identify non-convergence would be preferred.
 
-- it is difficult that there is not at single point at which convergence is reached: the algorithm will produce some fluctuations even after convergence [it's convergence in distribution, not to a point]  <!-- (and a ...
-distribution may not exist because of univariate modeling of multivariate distribution???) -->
+<!-- this may be inadequate because of reasons 1, 2, and 3 [see original text] -->
+## new alinea
 
-- therefore, we should look at *non*-convergence
+- it is challenging to arrive at a single point at which conv has been reached. the algo may prod some fluc even after it has conv. the aim is to conv to a distr and not to a single point.
 
-- there are non-convergence identifiers for other iterative algorithms (MCMC), but it is not known whether these are appropriate for imputation algorithms as well
+<!-- - it is difficult that there is not at single point at which convergence is reached: the algorithm will produce some fluctuations even after convergence [it's convergence in distribution, not to a point]  <!-- (and a ...
+distribution may not exist because of univariate modeling of multivariate distribution???) --> -->
+
+- because of this property, it may be more desirable to focus on non-c.
+
+<!-- therefore, we should look at *non*-convergence --> [discussie: als we geen non-conv kunnen observeren, moeten we ons dan zorgen maken om convergentie??] -->
+
+- fortunately there are non-convergence identifiers for other iterative algorithms. but the validity of these identifiers has not been syst eval on imp alg.
+<!-- (MCMC), but it is not known whether these are appropriate for imputation algorithms as well -->
 
@@
 
-- in this paper, we explore different methods for identifying non-convergence in iterative imputation algorithms
+- in this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
+we eval whether these are able to cover the extend of the non-c but we also invest the relation between non-c and the validity of the inf.
+
+we investigate whether these identifiers are able to
 
@@
 
-- we show how all this works by means of a case study
+- we demonst the workings of all this by means of a case study
 

### 2020-11-26 Manuscript.Rmd
Remove old into section

--- 2020-11-19 Manuscript.Rmd
@@
 bibliography: ref.bib
+abstract: "Iterative imputation has become the de facto standard to accommodate for the ubiquitous problem of missing data.
+While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
+Our study provides insight into identifying non-convergence in iterative imputation algorithms.
+We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
+In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
+We conclude that it never hurts to iterate longer, but such calculations hardly bring added value."
+
 ---
@@
 
-*Notation in this manuscript: *
+# Notation in this manuscript:
 
@@
 
-# Abstract
-
-Iterative imputation has become the de facto standard approach to accommodate the ubiquitous problem of missing data.
-While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
-Our study provides insight into identifying non-convergence in iterative imputation algorithms.
-We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
-In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
-We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
+- asterisks denote a weak segway between sentences
 
@@
 
-## One
+Iterative imputation has become the de facto standard to accommodate for missing data.
+The aim is usually to draw valid inferences, i.e.
+to get unbiased, confidence-valid estimates that incorporate the effects of the missingness.
+Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem.
+The missing values are imputed (i.e., filled in) using some sort of algorithm.
+And subsequently, the scientific model of interest is performed on the completed data.
+To obtain valid scientific estimates, both the missing data problem and the scientific problem should be considered appropriately.
+The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.
 
-- iterative imputation has become the de facto standard approach to accommodate missing data
+All inferences rely on convergence of the algorithm, **yet** determining whether an algorithm has converged is not trivial.* There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
+A widely accepted practice is visual inspection of the algorithm, **but** identifying convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
+Therefore, a quantitative diagnostic method to identify convergence would be preferred.
 
-- the aim is to draw valid inference: to get unbiased, confidence-valid estimates that incorporate the effect of the missingness
+It is challenging to arrive upon a single point at which convergence has been reached.* Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
+Because of this property, it may be more desirable to focus on *non*-convergence.
+Fortunately, there are non-convergence identifiers for other iterative algorithms, **although** the validity of these identifiers has not been systematically evaluated on imputation algorithms.
 
-- the estimates are obtained by separating the missing data problem from the scientific problem
-
-- the missing data problem is solved first: by imputing (i.e., filling in) the missing values in the incomplete dataset, using an algorithm such as MICE
-
-- then the analysis of scientific interest can be performed on the completed data, which yields the scientific estimates
-
-- to obtain valid scientific estimates, both the missing data problem and the scientific problem should be solved correctly
-
-- one aspect that is often overlooked is the convergence of the algorithm that is used to generate imputations
-
-## Two
-
-- there is no consensus on the convergence properties of iterative imputation algorithms, and there has not been a systematic study on how to evaluate the convergence of these algorithms
-
-- yet, the inferences rely on algorithmic convergence
-
-- the current practice is visual inspection, but this may be inadequate because of reasons 1, 2, and 3 [see original text]
-
-- it is difficult that there is not at single point at which convergence is reached: the algorithm will produce some fluctuations even after convergence [it's convergence in distribution, not to a point]  <!-- (and a ...
-distribution may not exist because of univariate modeling of multivariate distribution???) -->
-
-- therefore, we should look at *non*-convergence
-
-- there are non-convergence identifiers for other iterative algorithms (MCMC), but it is not known whether these are appropriate for imputation algorithms as well
-
-## Three
-
-- in this paper, we explore different methods for identifying non-convergence in iterative imputation algorithms
-
-- since the ultimate goal is to have inferential validity, we use this as performance measure in a simulation study: if the estimates are unbiased and confidence-valid, the non-convergence is, apparently, not influential
-
-- we develop guidelines for practice to interpret the non-convergence identifiers, because we typically don't have the ground truth (to calculate bias and coverage)
-
-- we show how all this works by means of a case study
-
-<!-- Iterative imputation has become the de facto standard approach to accommodate missing data problems.
-With iterative imputation, we 'impute' (i.e., fill in) plausible values in an incomplete dataset.
-These plausible values are generated using algorithms such as 'MICE' [@mice].
-These algorithms sequence through the variables in an incomplete dataset  -->
-
-<!-- Iterative imputation entails using algorithms such as MICE [@mice] to replace missing values by plausible values.
-These plausible values are sampled from the distribution of each missing entry, and draw a plausible value from this distribution to replace the missing value by.
--->
-
-<!-- While it is widely accepted that the technique can yield valid inferences under many different missing data circumstances, these inferences all rely on algorithmic convergence.
-We fill in the missing data by modeling a multivariate distribution for each missing entry.
-We update dis model by sequencing through the variables See Figure 1.
-Van Buuren et al.
-(2006) have shown that just a few iterations maybe sufficient.
-Most imputation software packages use a default value of five or ten iterations.
-But how can we know did this is sufficient?
--->
-<!-- There is currently no consensus on how to evaluate the convergence properties of iterative imputation algorithms.
-Imputation algorithms such as mice are a type of Gibbs sampler.
-It would does make sense to evaluate the convergence of iterative imputation algorithms using common MSMC non convergence identifiers.
-This has not been done.
--->
-<!-- With MCMC algorithms there is not a single point at which convergence is reached.  -->
-
-## Original text [remove later]
-
-Iterative imputation has become the de facto standard approach to accommodate missing data.
-With iterative imputation, we 'impute' (i.e., fill in) plausible data values in the place of missing values in an incomplete dataset.
-We obtain these plausible values by sampling from a distribution that we model for each missing entry and update sequentially.
-@buur06 have shown that with an iterative imputation algorithm such as 'MICE', just a few iterations are enough to yield unbiased, confidence-valid statistical inferences.
-
-<!-- Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences.
-The de facto approach to accommodate missing data is to 'impute' (i.e., fill in) any missing entries in an incomplete data.
-With iterative imputation, we model the distribution of each missing entry, and draw one or more plausible values from it.
--->
-
-<!-- With iterative imputation, we 'impute' (i.e., fill in) the missing data in an incomplete dataset by drawing values from the distribution of each missing entry.
-To generate imputed values, most imputation software packages rely on iterative algorithms.
-[Why iterative?
-We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.]  -->
-
-Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
-
-With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration.
-This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
-Are the imputations then to be trusted?
-And can we rely on the inference obtained using the imputed data?
-These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-[Common convergence diagnostics may not work.] While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@liu14; @zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence.
-<!-- (Van Buuren, 2018, 6.5.2) (Zhu & Raghunathan, 2015; Takahashi, 2017) -->
-
-Identifying non-convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence (Van Buuren, 2018, 6.5.2).
-Therefore, a quantitative diagnostic method to identify non-convergence would be preferred.
-
-In this paper, we explore some existing and new means of diagnosing non-convergence in iterative algorithms.
-We evaluate the validity of these methods in the context of missing data imputation using model-based simulation in `R` [@R].
-[Brief overview of paper: the diagnostics, the simulation, the results, the discussion, a case study.]
-
-<!-- In this paper, we explore diagnostic methods for iterative imputation algorithms.
-For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
-We consider two non-convergence identifiers for iterative algorithms: autocorrelation (conform Lynch, 2007, p.
-147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p.
-5).
-Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm.
--->
-
-<!-- We propose a novel multivariate parameter to check for non-convergence in iterative algorithms.
-We aim to show which method is the most informative about non-convergence in iterative imputation [explain that method = parameter + diagnostic + interpretation].
--->
-
-<!-- - Missing data has to be accommodated, usually filled in -->
-
-<!-- - Filling in is algorithmic, which relies on convergence -->
-
-<!-- - Diagnosing convergence is difficult, both visually and diagnostically -->
-
-<!-- - In this paper we explore diagnostic methods for MICE, and show which method (parameter + diagnostic + threshold) is the most informative about convergence  -->
-
-# Identifying non-convergence [remove later]
-
-- Iterative imputation is a sort of MCMC algorithm, so it makes sense to investigate non-convergence in a similar fashion to typical MCMC algorithms.
-
-- With MCMC, the generated values will vary even after convergence, so there is not a unique point at which convergence is established [@gelm13].
-Therefore, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
-Non-convergence occurs when one of two requirements for convergence is not met.
-
-- There are two requirements for convergence of iterative algorithms: mixing and
-stationarity [@gelm13]. Mixing implies that generated values intermingle nicely, and stationarity is characterized by the absence of trending between successive draws.
-
-- What are the consequences? --\> bias, under-coverage [refer to van Buuren (2018) instead of reproducing this]
-
-<!-- -   Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g.
-N\_eff is estimated using the variance-covariance matrix of the set of samples, while this is not the case in iterative imputation per se.
--->
-
-- We consider autocorrelation and rhat, and monitor four parameters: chain means, chain variances, a scientific estimate, and lambda.
-
-- Explain lambda
-
-```{r eval=FALSE, include=FALSE}
-# # maybe add ESS instead of AC???
-# LaplacesDemon::ESS(parameters[parameters$m == 1 & parameters$sim == 1, "qhat"])
-```
+In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
+We evaluate whether these methods are able to cover the extent of the non-convergence, **but** we also investigate the relation between non-convergence and the validity of the inferences.
+We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.
 
@@
 
-- the parameter can be any statistic that we track across iterations, for example chain means (i.e, the average imputed value per imputation)
+- the parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means)
 

### 2020-12-02 Manuscript.Rmd

@@
 
-- asterisks denote a weak segway between sentences
+- asterisks denote a weak 'segway' between sentences
 
@@
 And subsequently, the scientific model of interest is performed on the completed data.
-To obtain valid scientific estimates, both the missing data problem and the scientific problem should be considered appropriately.
+To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered.
 The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.
 
-All inferences rely on convergence of the algorithm, **yet** determining whether an algorithm has converged is not trivial.* There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
-A widely accepted practice is visual inspection of the algorithm, **but** identifying convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
-Therefore, a quantitative diagnostic method to identify convergence would be preferred.
+All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial.
+There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
+A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
+Therefore, a quantitative diagnostic method to assess convergence would be preferred.*
 
@@
 Because of this property, it may be more desirable to focus on *non*-convergence.
-Fortunately, there are non-convergence identifiers for other iterative algorithms, **although** the validity of these identifiers has not been systematically evaluated on imputation algorithms.
+Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
 
 In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
-We evaluate whether these methods are able to cover the extent of the non-convergence, **but** we also investigate the relation between non-convergence and the validity of the inferences.
+We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences.
 We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.
@@
 # Simulation Set-Up
-
-<!-- The aim of the simulation study is to determine the impact of non-convergence on the validity of statistical inferences, and to assess whether non-convergence may be detected using several diagnostic methods.
-Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions (nsim = 1000).
--->
 
@@
 We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.
-<!-- : three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??]. -->
 
@@
 
-## Methods
+## Diagnostic methods
 
@@
 
-As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2)[or just of the regression estimate??
-otherwise add the ciw and cov of r2 too!!].
+As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2).
+<!-- [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!]. -->
 
@@
 Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
-<!-- The next four plots show the autocorrelations in each parameter plotted against the number of iterations. -->
 
@@
 Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
-<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->
 
@@
 
-<!-- Update with different CI computation, see <https://stefvanbuuren.name/fimd/sec-evaluation.html#sec:quantifyingbias> -->
-
 Nominal coverage is quickly reached (see \@ref(fig:perf)C).
@@
 Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
-
-<!-- Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=6. -->
 
@@
 - Determining non-stationarity with lambda is more difficult than with qhat :(
-
-<!-- # Check if heuristic for AC works -->
-
-<!-- ```{r} -->
-
-<!-- load("Data/diagnostics.Rdata") -->
-
-<!-- load("Data/raw_outcomes.Rdata") -->
-
-<!-- sim_nr <- 3 -->
-
-<!-- outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line() -->
-
-<!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line() -->
-
-<!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line() -->
-
-<!-- ``` -->
 

### 2020-12-02 Non-converging-case-study.Rmd

## Case Study

- We use real data: the `boys` dataset from the `mice` package

- We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`

- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.

- The figures show results of a `mice` run with 30 iterations but otherwise default settings.

```{r echo=FALSE}
# autocorrelation plots (not used)
# casestudy_plot(v="ac.max.qhat") #+ list(scale_y_continuous(limits = c(0.65,1)))
# casestudy_plot(v="ac.max.lambda")
```

```{r echo=FALSE}
p <- plot(mids)
p[1:2,1] # %>% plotly::ggplotly()
```

From the traceplot of the chain means it seems that mixing improves up-to 20 iterations, and there is some trending up-to 20 iterations as well. The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations.

```{r echo=FALSE}
# p <- casestudy_plot(v="est", name = "Regression estimate")
# p %>% plotly::ggplotly()
```

```{r echo=FALSE}
p <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
  list(
    geom_hline(yintercept = 1.2, color = "grey")#, 
    #scale_y_continuous(limits = c(1,5))
    )
p # %>% plotly::ggplotly()
```

This figure shows that 36 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.

```{r echo=FALSE}
p <-
  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
  list(geom_hline(yintercept = 1.2, color = "grey")#,
       #scale_y_continuous(limits = c(1, 2.5))
       )
p # %>% plotly::ggplotly()
```

The scientific estimate reaches the threshold much sooner, when $it=15$.

```{r echo=FALSE}
p <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
  list(geom_hline(yintercept = 1.2, color = "grey")#,
       #scale_y_continuous(limits = c(1, 2.5))
       )
p # %>% plotly::ggplotly()
```

According to the $\widehat{R}$-values of the novel parameter, at least 25 iterations are required.

[Is this a nice example? Why is convergence so slow? Don't the defaults work? Is this the bmi+wgt+wgt problem? Maybe find an easier problem?]

### 2021-09-17 Manuscript.Rmd
Add reference against visual inspection of several parameters
Save manuscript before editing

@@
 
-Iterative imputation has become the de facto standard to accommodate for missing data.
-The aim is usually to draw valid inferences, i.e.
-to get unbiased, confidence-valid estimates that incorporate the effects of the missingness.
-Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem.
-The missing values are imputed (i.e., filled in) using some sort of algorithm.
-And subsequently, the scientific model of interest is performed on the completed data.
+Iterative imputation has become the de facto standard to accommodate missing data in social scientific research.
+The technique enables researchers to draw valid inferences despite having incomplete observations.
+Unbiased, confidence-valid estimates are obtained by separating the missing data problem from the scientific problem.
+First, the missing values are imputed (i.e., filled in) using some sort of algorithm.
+Only subsequently, the scientific model of interest is performed on the completed data.
 To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered.
-The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.
+The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.
 
-All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial.
+The scientific inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial.
 There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
@@
 We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.
+
+Add reference: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p.
+285).
 

### 2021-09-17 Manuscript.Rmd
re-order paragraphs in intro

@@
 
-Iterative imputation has become the de facto standard to accommodate missing data in social scientific research.
-The technique enables researchers to draw valid inferences despite having incomplete observations.
-Unbiased, confidence-valid estimates are obtained by separating the missing data problem from the scientific problem.
-First, the missing values are imputed (i.e., filled in) using some sort of algorithm.
-Only subsequently, the scientific model of interest is performed on the completed data.
-To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered.
+Iterative imputation has become the de facto standard to accommodate missing data in social scientific research [ref: e.g., rubin after 18 years paper?
+or murray 2018?].
+The technique enables researchers to draw valid inferences when faced with [phrasing: despite of?] incomplete observations.
+Every missing cell in an incomplete dataset is imputed (i.e., filled in) using an imputation algorithm.
+And once the dataset is completed, an analysis of scientific interest can be performed.
+This analysis will yield unbiased and confidence-valid estimates when both the missing data problem and the scientific problem are appropriately considered.
 The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.
+Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
 
-The scientific inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial.
-There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
-A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
-Therefore, a quantitative diagnostic method to assess convergence would be preferred.*
+<!-- Any scientific inferences obtained after iterative imputation rely on the convergence of the imputation algorithm,  -->
 
-It is challenging to arrive upon a single point at which convergence has been reached.* Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
+Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation.* It is challenging to arrive upon a single point at which convergence has been reached.
+Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
 Because of this property, it may be more desirable to focus on *non*-convergence.
+A widely accepted practice is visual inspection of the algorithm [@ragh07; @buur18].
+Diagnosing non-convergence through visual inspection may, however, be undesirable: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
+[add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p.
+285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred.
 Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
@@
 We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.
-
-Add reference: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p.
-285).
 
@@
 Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
-To induce non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
+To assess non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
 We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.

### 2021-09-30 Manuscript.Rmd
Add missingness conditions figure

@@
 bibliography: ref.bib
-abstract: "Iterative imputation has become the de facto standard to accommodate for the ubiquitous problem of missing data.
+abstract: "Iterative imputation has become the de facto standard to accommodate the ubiquitous problem of missing data.
 While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence.
@@
 Iterative imputation has become the de facto standard to accommodate missing data in social scientific research [ref: e.g., rubin after 18 years paper?
-or murray 2018?].
-The technique enables researchers to draw valid inferences when faced with [phrasing: despite of?] incomplete observations.
-Every missing cell in an incomplete dataset is imputed (i.e., filled in) using an imputation algorithm.
-And once the dataset is completed, an analysis of scientific interest can be performed.
-This analysis will yield unbiased and confidence-valid estimates when both the missing data problem and the scientific problem are appropriately considered.
+murray 2018?].
+The technique enables researchers to draw valid inferences when faced with incomplete observations, without resorting to ad hoc solutions such as list-wise deletion which may bias results.
+With iterative imputation, every missing cell in an incomplete dataset is imputed (i.e., filled in) algorithmically.
+And once the dataset is completed, the analysis of scientific interest can be performed.
+This analysis will yield unbiased and confidence-valid estimates if--and only if--both the missing data problem and the scientific problem are appropriately considered.
 The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.
@@
 
-Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation.* It is challenging to arrive upon a single point at which convergence has been reached.
+Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation.
 Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
@@
 285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred.
-Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
+Fortunately, there are non-convergence identifiers for *other* iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
+
+<!-- It is challenging to arrive upon a single point at which convergence has been reached. -->
 
@@
 Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
-To assess non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
-We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.
+To assess non-convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
+We differentiate between nine different missingness scenarios that are defined in the data generating mechanism.
 
@@
 
-Data are generated in each simulation repetition for a complete set of $n_{obs} = 1000$ cases (i.e., before inducing missingness). We define three multivariately normal random variables, let
+We generate one complete set of $n_{obs} = 1000$ cases in each simulation repetition. This set consists of three multivariately normal random variables, such that
 
@@
 
-The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases.
+The complete set is 'amputed' (i.e., made incomplete) according to nine missingness conditions.
+We use a $3 \times 3$ factorial design with three missingness mechanisms and three proportions of incomplete cases, see Figure XYZ.
+Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
+There are three distinct types of mechanisms, as defined by Rubin (1976).
+Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), the probability of being missing depends on observed information (MAR; Missing At Random), or the probability depends on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).
+[add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018].
 
-- we use the three missingness mechanisms MCAR, MAR, and MNAR with defaults settings from `mice::ampute()` (e.g., right-tailed MAR)
+```{r figure-missingness, echo=FALSE}
+tibble(prop = c(25,50,75), MCAR = 1:3, MAR = 4:6, MNAR = 7:9) %>% 
+  pivot_longer(cols = c(MCAR, MAR, MNAR)) %>% 
+  mutate(prop = factor(prop, levels = c(25, 50, 75), labels = c("25%", "50%", "75%")),
+         mech = factor(name, levels = c("MCAR", "MAR", "MNAR")),
+         text = paste(mech, "missingness\n", prop, "incomplete")) %>% 
+  ggplot(aes(x = mech, y = prop, fill = as.factor(value), label = text)) +
+  geom_tile(color = "black") +
+  geom_text() +
+  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
+  scale_y_discrete(limits = c("75%", "50%", "25%")) +
+  scale_x_discrete(breaks = c("MCAR", "MAR", "MNAR")) +
+  theme_minimal() +
+  labs(x = "Missingness mechanism", y = "Proportion of incomplete cases")
+```
 
-- we use a multivariate missing data pattern and make 25%, 50%, and 75% of cases incomplete
+<!-- - we use the three missingness mechanisms MCAR, MAR, and MNAR with defaults settings from `mice::ampute()` (e.g., right-tailed MAR) -->
+
+<!-- - we use a multivariate missing data pattern and make 25%, 50%, and 75% of cases incomplete -->
 
@@
 We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice].
-On each imputed dataset, we perform multiple linear regression to predict outcome variable $Y$ from the other two variables
+On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
+We predict outcome variable $Y$ from the other two variables,
 
@@
 
-As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2).
+As recommended by van Buuren, our performance measures are bias, confidence interval width, and coverage rate of the estimands (-@buur18 $\S$ 2.5.2).
+These measures are averaged across the simulation repetitions.
 <!-- [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!]. -->
